---
title: Three Measures of Inequality (Making Of)
author: Jose M Sallan
date: '2023-05-08'
slug: hree-measures-of-inequality-making-of
categories:
  - R
  - economics
tags:
  - data visualization
  - distributions
  - dplyr
  - ggplot
meta_img: images/image.png
description: Description for the page
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

This post is the making of [the previous post about measures of inequality](https://jmsallan.netlify.app/blog/three-measures-of-inequality/). I will present how I have:

* defined the three income distributions.
* obtained the mean to median ratio for each distribution.
* defined an efficient function to calculate the Gini index.
* obtained the Lorenz curve plots.

I have used the `tidyverse` to do all the jobs and `kableExtra` to present tables.

```{r}
library(tidyverse)
library(kableExtra)
```

## Defining Income Distributions

I start defining an `income` table with three distributions progressively more unequal. `r1` is a sample of a normal distribution, and `r2` and `r3` are samples of right skewed distributions obtaining powering 10 with the values of a normal distribution. The standard deviation of `r3` is higher than `r2`, resulting in higher inequality.

```{r}
set.seed(1111)
income <- tibble(id = 1:10000,
                 r1 = rnorm(10000, mean = 10000, sd = 100),
                 r2 = 10^rnorm(10000, mean = 4, sd = 0.2),
                 r3 = 10^rnorm(10000, mean = 4, sd = 0.5))
```

Let's define an `income_long` tidy table with `pivot_longer`:

```{r}
income_long <- income |>
  pivot_longer(-id)
income_long
```

## Plotting the Histograms

The obvious choice to plot the three histograms is to use facet_grid:

```{r, out.width='100%'}
income_long |>
  ggplot(aes(value)) +
  geom_histogram(bins = 40, fill = "#A0A0A0") +
      theme_minimal() +
      labs(x = NULL, y = NULL) +
      facet_grid(. ~ name)
```

This first draft is not adequate as the ranges of values for each distribution are different. To remedy that, I have set `scale = "free"` argument in `facet_grid`. I have also changed the size and angle of x axis labels to avoid overlap.

```{r}
income_long |>
  ggplot(aes(value)) +
  geom_histogram(bins = 40, fill = "#A0A0A0") +
      facet_grid(. ~ name, scales = "free") +
      theme_minimal() +
      theme(axis.title = element_blank(), 
            axis.text.x = element_text(size = 10, angle = -45))
```

## Mean to median Ratio

To obtain the mean to median ratio I need to `summarise` the mean and median, and then obtain the ratio with `mutate`. I can adjust the decimals to show with `kbl`:

```{r}
income_long |>
  group_by(name) |>
  summarise(mean = mean(value), median = median(value)) |>
  mutate(ratio = mean/median) |>
  kbl(digits = 3) |>
  kable_styling(full_width = FALSE)
```

## Calculating the Gini Index

I need to define a function to implement this expression.

\[\frac{\displaystyle\sum_{i=1}^n \displaystyle\sum_{i=1}^n \vert x_i - x_j \vert}{2n^2\bar{x}} \]

The obvious path is the one defined in the `gini2` function, looping twice across each element of the vector:

```{r}
gini2 <- function(x){
  n <- length(x)
  sum <- 0
  
  for(i in 1:n)
    for(j in 1:n)
      sum <- sum + abs(x[i] - x[j])
  
  g <- sum / (2 * n^2 * mean(x))
  return(g)
}
```

But as R is a **vectorial language**, we can try an alternative to find all absolute differences without looping. I will define two square matrices: one with all rows equal and other with all columns equal. We can use the rep function to do this. We obtain the first matrix doing:

```{r}
matrix(rep(1:10, each = 10), 10, 10)
```

And the second:

```{r}
matrix(rep(1:10, times = 10), 10, 10)
```

So we obtain all differences doing:

```{r}
abs(matrix(rep(1:10, each = 10), 10, 10) - matrix(rep(1:10, times = 10), 10, 10))
```

And the summmation:

```{r}
sum(abs(matrix(rep(1:10, each = 10), 10, 10) - matrix(rep(1:10, times = 10), 10, 10)))
```

We can obtain the same result substracting the vectors used to build the matrices:

```{r}
sum(abs(rep(1:10, each = 10) - rep(1:10, times = 10)))
```

So we can construct the function:

```{r}
gini <- function(x){
  n <- length(x)
  g <- sum(abs(rep(x, times = n) - rep(x, each = n))) / (2 * n^2 * mean(x))
  return(g)
}
```

Let's examine the performance of each function with `rbenchmark`:

```{r}
rbenchmark::benchmark(replications = 10,
                      gini(income$r3),
                      gini2(income$r3),
                      columns=c('test', 'elapsed', 'relative', 'replications'),
                      order = 'elapsed')
```

The non-looping function has better performance, so I will use `gini` to calculate gini indices of each distribution:

```{r}
income |>
  summarise(across(r1:r3, gini)) |>
  kbl(digits = 3) |>
  kable_styling(full_width = FALSE)
```


## Plotting Lorenz Curves

The last job is plotting Lorenz curves. To do so I need to calculate:

* the cumulative share of people from lower to higher income.
* he cumulative share of income earned by each share.

This can be achieved with `dplyr` functions:

* `arrange` rows by `name` and `value` of income in decreasing order.
* obtain the cumulative fraction of individuals `frac_ind`.
* obtain the cumulative fraction of income using `cumsum`.

```{r}
income_long <- 
  income_long |>
  arrange(name, value) |>
  group_by(name) |>
  mutate(frac_ind =1:n()/n(),
         frac_rent = cumsum(value)/sum(value), .groups = "drop")
```

Note that here I am combining `group_by` and `mutate`. As I am storing the result in income_long and I don't need a grouping table, I am doing `.groups = "drop"` in `mutate`.

Now I am ready to plot the Lorenz curves. The area under the curve is filled with `geom_polygon` quite naturally in this case. For this plot, I have decided to use a pale yellow background to differentiate each curve visually adding a `panel.background` definition in `theme`.

```{r, out.width='100%'}
income_long |>
  ggplot(aes(frac_ind, frac_rent)) +
  geom_line() +
  geom_polygon(fill = "#A0A0A0") +
  theme_minimal() +
  theme(panel.background = element_rect(fill = "#FFFFEE")) +
  labs(x=NULL, y=NULL) +
  facet_grid(. ~ name)
```

## Session Info

```{r, echo=FALSE}
sessionInfo()
```


