---
title: Mapping with purrr
author: Jose M Sallan
date: '2022-11-30'
slug: mapping-with-purrr
categories:
  - R
tags:
  - purrr
  - R
meta_img: images/image.png
description: Description for the page
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

In this post, I will present the functionalities of the `purrr` package for mapping (iterating) along vectors, lists or data frames using functional programming. `purrr` is included in the tidyverse and it is loaded with the `tidyverse` (meta-)package.

I will also be using `kableExtra` to present data frames nicely.

```{r}
library(tidyverse)
library(kableExtra)
```

## Mapping one list

In mathematics, **mapping** is an operation that associates each element of a given set (the domain) with one or more elements of a second set (the range). This is precisely what the `map` function of `purrr` is doing. 

`map` has two arguments. The first one is a list, vector or data frame. The second is the function to be applied to each element of the first argument. 

Let's build a **list** of a vector and two data frames:

```{r}
l <- list(a = LETTERS, b = iris, c = mtcars)
```

The outcome of `map` is always a list:

```{r}
map(l, length)
```

We can obtain a similar output using the `lapply` R base function:

```{r}
lapply(l, length)
```

The `length` function always returns an integer, so it makes sense to obtain a vector of integers instead of a list. We can achieve that with `map_int`.

```{r}
map_int(l, length)
```

Again, we can obtain a similar output using `sapply`:

```{r}
sapply(l, length)
```

All mapping functions of `purrr` include variants that allow specifying the class of the output. Using `map_int`, `map_dbl`, `map_chr` and `map_lgl` we obtain, if possible, outputs of class integer, double, character and logical respectively.

Let's obtain the output of `length` for list `l` as a character vector:

```{r}
map_chr(l, length)
```

The first argument of the `map` family functions can also be a data frame. Those functions treat a data frame as a list of columns. Let's see how can we calculate the mean of each of the columns of `mtcars`.

```{r}
map_dbl(mtcars, mean)
```

Using R base we can obtain the same result using `apply` across columns:

```{r}
apply(mtcars, 2, mean)
```

In `purrr` functions we can use **function shortcuts**, where the function is introduced with `~`. In `map`, the element of list or data frame is represented as `.`.

```{r}
map_dbl(mtcars, ~ round(mean(.), 4))
```

We can produce a similar output using R base, but not with the function shortcut.

```{r}
sapply(mtcars, \(i) round(mean(i), 4))
```

## Mapping two lists

The `map2` family of functions allows iterating a function with two arguments along two lists.

To illustrate how `map2` functions work, let's build a function that tells us if we have improved or worsened our performance when comparing `past` and `present` grades:

```{r}
check_improvement <- function(past, present){
  
  if(past < present){
    report <- "improved"
  }else{
    report <- "not improved"
  }
  
  return(report)
}
```

We want to apply `check_improvement` to two vectors of past and present grades:

```{r}
set.seed(1111)
past_grades <- sample(1:10, 10, replace = TRUE)
present_grades <- sample(1:10, 10, replace = TRUE)
```

We cannot apply `check_improvement` to the vectors `past_grades` and `present_grades` directly, as `if` only logical evaluations arguments of length one. We can iterate along these two vectors using `map2`. The two inputs of the function are labeled as ´.x` and `.y`.

```{r}
map2(past_grades, present_grades, ~ check_improvement(.x, .y))
```

Using functions `*_dfr` and `*_dfc` we can present the output as tibbles constructed by rows or columns, respectively. Let's modify the function above to return a row of a data frame for each observation.

```{r}
check_improvement2 <- function(past, present){
  
  if(past < present){
    report <- "improved"
  }else{
    report <- "not improved"
  }
  
  return(list(past = past, present = present, report = report))
}
```

Now we get the data frame binding rows with `map2_dfr`:

```{r}
map2_dfr(past_grades, present_grades, ~ check_improvement2(.x, .y)) %>%
  kbl() %>%
  kable_styling(full_width = FALSE)
```

## Mapping more than two lists

We can map functions taking three or more arguments using the `pmap` family. They work similary to `map2`, but taking arguments of the form `..1`, `..2`, `..3` and so on. The input of those functions is a list with the elements to iterate. Let's see how the `pmap` functions work with an example.

Let's consider a quiz where you are betting on the results of football matches. If your result has the same winning team as the real match, or you correctly guess a tie, you get two points. If your bet matches the exact result, you get three points. If `f1` and `f2` are the forecasted goals for each team, and `r1` and `r2` the real result, we can get the points of the bet with the function:

```{r}
score <- function(f1, f2, r1, r2){
  
  points <- 0
  
  if(sign(f1-f2) == sign(r1-r2))
    points <- 2
  
  if(f1 == r1 & f2 == r2)
    points <- 3
  
  return(points)
}
```

Let's test the function with a list of bets and results `matches_list` and the `pmap_dbl` function:

```{r}
matches_list <- list(mf1 = c(0, 0, 0, 1),
                     mf2 = c(0, 2, 3, 1),
                     mr1 = c(1, 0, 1, 1),
                     mr2 = c(1, 2, 0, 1))

pmap_dbl(matches_list, ~score(..1, ..2, ..3, ..4))
```

## Mapping in a data frame

We can use the `map`, `map2` and `pmap` families of functions inside a data frame or tibble using `mutate`. For the example above we can store bets and results in a tibble, and then use `mutate` to add a column with the results of the `score` function. Note that the argument of `pmap` can be also a data frame.

```{r}
matches <- tibble(mf1 = c(0, 0, 0, 1),
                  mf2 = c(0, 2, 3, 1),
                  mr1 = c(1, 0, 1, 1),
                  mr2 = c(1, 2, 0, 1))

matches <- matches %>%
  mutate(result = pmap_dbl(matches, ~score(..1, ..2, ..3, ..4)))

matches %>%
  kbl() %>%
  kable_styling(full_width = FALSE)
```

If we use functions of the `map` or `map2` family, the arguments of the functions are the columns of the data frame.

```{r}
grades <- tibble(past = past_grades, present = present_grades)

grades <- grades %>%
  mutate(check = map2_chr(past, present, ~ check_improvement(.x, .y)))

grades %>%
  kbl() %>%
  kable_styling(full_width = FALSE)
```


## References

* `purrr` page in tidyverse website <https://purrr.tidyverse.org/>
* *21: iteration* in Wickham, H. and Grolemund, G. (in progress). *R for data science*. <https://r4ds.had.co.nz/iteration.html>

## Session info

```{r, echo=FALSE}
sessionInfo()
```

