---
title: Examining correlation matrices in R
author: Jose M Sallan
date: '2021-12-27'
slug: examining-correlation-matrices-in-r
categories:
  - R
  - statistics
tags:
  - correlation
  - R
meta_img: images/image.png
description: Description for the page
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

In this post, I will introduce how to examine relationships between variables in a multivariate dataset using covariance and correlation matrices. Those matrices are the input of techniques like exploratory and confirmatory factor analysis and structural equation modelling. I this post, I will be using the `mtcars` dataset, that allows showing positive and negative relationships between variables.

In base R, we use `cov` and `cor` to obtain the covariances and correlation matrices of a multivariate distribution. These functions take a data frame with the observations as input.

```{r}
c_mat <- cov(mtcars)
r_mat <- cor(mtcars)
```

The **covariance matrix** $\mathbf{S}$ includes covariances $s_{ij}$ between all pairs of variables $\left(i, j\right)$ of the distribution. As $s_{ij} = s_{ji}$, it is a symmetric matrix. Diagonal elements $s_{ii}$ are the variance of each variable $i$.

```{r}
c_mat
```

Covariance values depend on the scale of each pair of variables and thus they are difficult to interpret. That's why we usually to examine the **correlation matrix** $\mathbf{R}$. Correlations $r_{ij}$ are scaled between -1 and +1, and diagonal elements are equal to one.

```{r}
r_mat
```

There are many R packages dealing with correlation matrices, to allow a better visualization and interpretation. Here I will present some functionalities of `corrr` and `corrplot` packages.

```{r}
library(corrr)
library(corrplot)
```

## The corrr package

The corrr package is a part of the tidymodels ecosystem, and allows manipulating and presenting correlation matrices as data frames. We use `correlate` to obtain correlations with `corrr`.

```{r}
r_df <- correlate(mtcars)
```

The outcome of `correlate` is a tibble, instead of a matrix. Variable names of rows are stored in an additional `term` column. By default values of diagonal are set to `NA`.

```{r}
r_df
```

With `stretch` we can get correlations as a long table:

```{r}
stretch(r_df)
```

With `focus` we can examine a part of the correlation matrix. Columns are the second argument of the function, and rows the rest of variables:

```{r}
focus(r_df, c(mpg, cyl))
```

`fashion` allows a pretty presentation of the correlation matrix. We can specify the number of `decimals`, and select if we want to print the `leading_zeros`. Here I am presenting the default input.

```{r}
fashion(r_df)
```

To interpret a correlation matrix, it can be useful to change the default order of variables, putting together highly correlated variables. We accomplish this with the `rearrange` function. The methods available to rearrange variables are principal components analysis `"PCA"` (the default) or hierarchical clustering `"HC"`.

```{r}
rearrange(r_df, method = "HC")
```

Correlation matrices are symmetric and with ones in the diagonal, so it is frequent to present its lower triangular part without diagonal elements. We get this with `shave`:

```{r}
shave(r_df)
```

We can achieve a more satisfying presentation combining `shave` and `fashion`:

```{r}
fashion(shave(r_df))
```

We can also plot the correlation matrix with `rplot`. It is customary to `rearrange` and `shave` the matrix before plotting:

```{r, fig.align='center', out.width='100%'}
r_pretty <- shave(rearrange(r_df))
rplot(r_pretty)
```

## The corrplot package

`corrplot` provides a visual exploratory tool of correlation matrices that supports automatic variable reordering to help detect hidden patterns among variables. It can be seen as a visual alternative to exploratory factor analysis.

The functionalities of corrplot are nicely explained in the [package vignette](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html). Here I will be posting some illustrative examples.

We specify how to present correlations with the `method` argument of the `corrplot` function:

```{r, fig.align='center', out.width='100%'}
corrplot(r_mat, method = 'number') 
```

We can specify how to `order` variables in the correlation matrix. Methods available are angular order of eigenvectors `"AOE"`, principal components `"FPC"` and hierarchical clustering `"HC"`.

```{r, fig.align='center', out.width='100%'}
corrplot(r_mat, method = "circle", order = "hclust", diag = FALSE)
```

The `corrplot.mixed` function allows presenting two different visualizations of the same correlation matrix in the upper and lower triangular parts of the matrix.

```{r, fig.align='center', out.width='100%'}
corrplot.mixed(r_mat, upper = 'ellipse', lower = "shade", order = "hclust")
```

## Examining covariance and correlation matrices in R

Covariance and correlation matrices express relationships between variables of a multivariate sample. As correlations are scaled between -1 nd +1, it is more convenient for humans to examine correlation matrices. With `corrr` and `corrplot` packages we can examine correlation matrices, group highly correlated subsets of variables and present visualizations of the results.

After examining correlation matrices, we can engage in advanced techniques to examine correlational structures, like exploratory and confirmatory factor analysis or structural equation modelling.

## References

* Covariance and Pearson correlation in R <https://jmsallan.netlify.app/blog/covariance-and-pearson-correlation-in-r/>
* `corrr` vignette: <https://corrr.tidymodels.org/>
* `corrplot` vignette: <https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html>

## Session info

```{r, echo=FALSE}
sessionInfo()
```

