---
title: 2-opt Moves in the Travelling Salesman Problem
author: Jose M Sallan
date: '2024-05-08'
slug: 2-opt-moves-in-the-travelling-salesman-problem
categories:
  - optimization
  - R
tags:
  - ggraph
  - optimization
  - purrr
meta_img: images/image.png
description: Description for the page
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r, message=FALSE}
library(purrr)
library(tidygraph)
library(ggraph)
library(gridExtra)
library(patchwork)
```

The **travelling salesman problem (TSP)** is a classical optimization problem in graphs consisting in finding the cycle that visits each node exactly once of minimal total distance. It is a classical problem in operations research and it has been tackled with many approaches, including local search algorithms.

A classical neighborhood definition to implement local search algorithms is the **2-opt** move. This move consists in picking a pair of non contiguous edges, and rearrange the connections between the four nodes of the two edges as presented in the following example.

```{r, out.width='100%', fig.height=3, echo = FALSE}
move_2opt <- function(vec, i, j, n){
  if(j == n)
    v <- vec[c(1:(i-1), j:i)]
  else
    v <- vec[c(1:(i-1), j:i, (j+1):n)]
  return(v)
}

n <- 8
l <- matrix(c(cos(2*pi*(0:(n-1))/n), sin(2*pi*(0:(n-1))/n)), ncol = 2)

t_edges <- data.frame(i = 1:n, j = c(2:n, 1), pos = c(rep("n", 2), "r", rep("n", 2), "r", rep("n", 2)))
g_t <- tbl_graph(edges = t_edges, directed = FALSE)
g_t <- g_t |> activate(nodes) |> mutate(name = 1:n)

sol <- move_2opt(1:8, 4, 6, 8)

u_edges <- data.frame(i = sol, j = sol[c(2:n, 1)],
                  pos = c(rep("n", 2), "r", rep("n", 2), "r", rep("n", 2)))

g_u <- tbl_graph(edges = u_edges, directed = FALSE)
g_u <- g_u |> activate(nodes) |> mutate(name = 1:n) 

p_graph <- 

g1 <- ggraph(g_t, layout = l) + 
  geom_node_text(aes(label = name), size = 4) +
  geom_edge_link(aes(color = pos), 
                 start_cap = circle(3, 'mm'), end_cap = circle(3, 'mm'), 
                 alpha = 0.5, linewidth = 1) +
  theme_graph() +
  scale_edge_color_manual(values = c("black", "red")) +
  theme(legend.position = "none")

g2 <- ggraph(g_u, layout = l) + 
  geom_node_text(aes(label = name), size = 4) +
  geom_edge_link(aes(color = pos), 
                 start_cap = circle(3, 'mm'), end_cap = circle(3, 'mm'), 
                 alpha = 0.5, linewidth = 1) +
  theme_graph() +
  scale_edge_color_manual(values = c("black", "red")) +
  theme(legend.position = "none")

g1 + g2
```

The cycle of the right can be obtained from the cycle on the left reversing the $\{4,5,6\}$ sequence, so we can label this move as $(4,6)$.

In this post, we will cover some topics related with the 2-opt move

-   How many different 2-opt moves exist in a TSP instance of size `n`?
-   How can we generate all possible 2-opt moves for a solution?
-   How can we calculate in a fast way the total distance of the cycle resulting from a 2-opt move?
-   How can we evaluate total distance for all 2-opt moves of a solution, and choose the move that best improves total distance.

Let's see how can ask these questions with the R programming language extended by the tidyverse. I will be using the following packages:

-   The `purrr` for iterating effectively.
-   The `tidygraph` and `ggraph` packages to handle and visualize network data.
-   The `gridExtra` and `patchwork` packages to present visualizations on a grid.

Except from purrr, the packages have been used to present figures of 2-opt moves. They have been loaded previously to allow plotting the above figure.

## Number of 2-opt Moves

To find the number of 2-opt possible moves, we need to consider that the edges involved in the node must have different nodes. For each of the $n$ edges of a cycle covering all nodes, only $n-3$ edges have this property, as we need to exclude the edge itself and the two edges sharing a node with the considered edge. This leads to $n(n-3)$ possible moves.

But if the problem is symmetric, each 2-opt move has another move yielding the same effect. For the plot above, the $(4, 6)$ move is equivalent to the $(3, 7)$ move. Therefore, the total number of 2-opt moves in a symmetric TSP instance of size $n$ is $n(n-3)/2$.

## Generating 2-opt Moves

The function `move_2opt_table()` generates all 2-opt moves for a symmetric TSP in the following way:

-   We generate the `edges` data frame with the positions of node edges `(i. j)` of a solution.
-   For each edge, I find all edges `e2` with no common nodes for each edge `e1`. Then, I obtain the `i` element as the second node of the first edge, and `j` as the first node of the second edge.
- Finally, I choose moves with `i > j` to pick those moves with sequences to revert not including the first node.

```{r}
move_2opt_table <- function(n){
  
  edges <- data.frame(edge = 1:n, i = 1:n, j = c(2:n, 1))
  e1 <- rep(1:n, each = n-3)
  e2 <- c(sapply(1:n, \(i) ((i+1):(i+n-3) %% n) + 1))
  
  moves <- data.frame(e1 = e1, e2 = e2, i = e1+1, j = e2)
  
  moves <- moves[which(moves$i < moves$j), 3:4]
  
  return(moves)
}
```

The function is encoded in R base, so there is no additional package to use it.

Here is the result for `n = 7`...

```{r}
move_2opt_table(7)
```

... and the result for `n = 8`:

```{r}
move_2opt_table(8)
```

## Obtanining 2-opt Moves

The moves selected by `move_2opt_table()` reverse sequences that do not contain the first node. This makes the coding of `move_2opt()` simpler, as we need to care only for the case where $j$ is equal to the $n$-eth node.

```{r}
move_2opt <- function(vec, i, j, n){
  if(j == n)
    v <- vec[c(1:(i-1), j:i)]
  else
    v <- vec[c(1:(i-1), j:i, (j+1):n)]
  return(v)
}
```

Here is the move $(4,6)$ of the solution $\{1,2,3,4,5,6,7,8\}$, presented in the figure above.

```{r}
move_2opt(1:8, 4, 6, 8)
```

We can use the `purrr::pmap()` function to iterate along the table of 2-opt moves and obtain all solutions as a list.

```{r}
n <- 6
moves_6 <- move_2opt_table(n)
pmap(moves_6, ~  move_2opt(vec = 1:n, .x, .y, n = n))
```

In the following figure, I am presenting the twenty 2-opt moves for a TSP instance of size eight.

```{r, out.width='100%', echo=FALSE, fig.height=10}
plot_move <- function(sol){
  
  n <- length(sol)
  l <- matrix(c(cos(2*pi*(0:(n-1))/n), sin(2*pi*(0:(n-1))/n)), ncol = 2)
  
  t <- data.frame(i = sol, j = sol[c(2:n, 1)])
  g <- tbl_graph(edges = t, directed = FALSE)
  g <- g |> activate(nodes) |> mutate(name = 1:n)

  p <- ggraph(g, layout = l) + 
  geom_node_text(aes(label = name), size = 3) +
  geom_edge_link(start_cap = circle(2, 'mm'), end_cap = circle(2, 'mm'), 
                 alpha = 0.5, linewidth = 1) +
  theme_graph()
  
  return(p)
}

n <- 8
moves <- move_2opt_table(n)
plots <- pmap(moves, ~ plot_move(move_2opt(vec = 1:n, .x, .y, n = n)))
grid.arrange(grobs = plots, ncol = 4, nrow = 5, byrow = TRUE)
```

## Value of Solution for 2-opt Moves

In a local search algorithm like simulated annealing or tabu search, we need to perform many evaluations of the total distance of a solution, so it is relevant to calculate them in a fast way. Let's illustrate how to do this with a distance matrix `D` of a TSP instance of size `n = 8`.

```{r}
points <- function(n){
  df <- data.frame(x = sample(1:(2*n), n),
                   y = sample(1:(2*n), n))
  return(df)
}

n <- 8
set.seed(1111)
D <- dist(points(n), method = "manhattan", upper = TRUE) |> as.matrix()
```

The TSP() function calculates the value of a solution in a fast way, because we are not iterating to obtain edge values.

```{r}
TSP <- function(v, sol, n) sum(v[sol[c(1:(n-1), 1)] + (sol[c(2:n, n)] - 1)*n])
```

If we look again at the first picture, we observe that when we perform a 2-opt move, we remove two edges and add two new ones. So substracting and adding edge distances can be used to calculate objective function faster.

```{r}
TSP_2opt <- function(D, fit, i, j, n){
  if(j == n)
    new_sol <- fit + D[i-1, j] + D[i, 1] - D[i-1, i] - D[j, 1]
  else
    new_sol <- fit +  D[i-1, j] + D[i, j+1] - D[i-1, i] - D[j, j+1]
  
  return(new_sol)
}
```

Let's check if the function works. We start calculating the total distance of a solution `sol`.

```{r}
sol <- 1:n
fit <- TSP(as.numeric(D), sol, 8)
fit
```

Now we generate a new solution with the $(3,6)$ move, and compare the result obtained in the two ways.

```{r}
new_sol <- move_2opt(1:n, 3, 6, 8)
TSP(as.numeric(D), new_sol, 8) 
TSP_2opt(D, fit, 3, 6, 8)
```

We observe that both functions lead to the same result.

## Exploring a 2-opt Neighbourhood

We can use `purrr:pmap()` on the table of 2-opt moves to evaluate the value of cycle of each move, and find the best move arranging the table.

```{r}
moves |>
  mutate(sol = pmap_dbl(moves, ~ TSP_2opt(D, fit, .x, .y, n))) |>
  arrange(sol, i)
```

In this case, we observe that the best move for this solution is $(4,6)$, which reduces total cycle value from 92 to 68.

## Session Info

```{r, echo=FALSE}
sessionInfo()
```
