---
title: "Merging with dtplyr"
author: Jose M Sallan
date: '2021-04-15'
slug: merging-with-dtplyr
categories:
  - R
tags:
  - R
  - data.table
  - dplyr
  - data structures
meta_img: images/image.png
description: Description for the page
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

[In a recent post](https://jmsallan.netlify.app/blog/merging-with-data-table/), I illustrated how to merge tables in R using the `data.table` package. This package handles large tabular data effectively, although some find its syntax somewhat arcane. That's why Hadley Wickham has developed `dtplyr` "a `data.table` backend for `dplyr`, allowing you to write `dplyr` code that is automatically translated to the equivalent `data.table` code."

For `dtplyr` to work, we need to load also `dplyr`, and it is convenient to load `data.table` as well:

```{r}
library(data.table)
library(dplyr)
library(dtplyr)
```

```{r, echo=FALSE}
table1 <- data.table(id = c("a", "a", "b", "b", "c", "e"), 
                     v = 1:6,
                     date = as.IDate(c("2021-01-01", "2021-03-23", "2021-02-13", "2021-05-28", "2021-12-17", "2021-12-31")))

table2 <- data.table(id = c("a", "b", "c", "d"), w = 1:4)

table3 <- data.table(id = c("a", "a", "a", "b", "c", "f"), g = 21:26)

table4 <- data.table(id = c(rep("a", 5), rep("b", 3), rep("c")), h = 31:39)

table5 <- data.table(id = c(rep("a", 5), rep("b", 3), rep("c")), 
                     h = 31:39,
                     date = as.IDate(c("2021-01-01", "2021-02-04", "2021-03-23", "2021-06-03", "2021-08-15", "2021-02-13", "2021-05-28", "2021-07-12", "2021-12-17")))
table1_tibble <- tibble(id = c("a", "a", "b", "b", "c", "e"), 
                     v = 1:6,
                     date = as.IDate(c("2021-01-01", "2021-03-23", "2021-02-13", "2021-05-28", "2021-12-17", "2021-12-31")))
table2_tibble <- tibble(id = c("a", "b", "c", "d"), w = 1:4)
```

Let's compare the syntax of `data.table` and `dplyr` to merge tables. Let's consider two data tables:

```{r}
table1
table2
```

## Inner join

The result of an *inner join* is a table with the rows with values of merging variables existing in both tables:

```{r}
setkey(table1, "id")
setkey(table2, "id")
merge(table1, table2)
```

In `dplyr` we use `inner_join` for that. If the input are two `data.table` objects and `dtplyr` is loaded, we perform the same operation as above. We use `as.data.table()` to obtain a `data.table` object.

```{r}
inner_join(table1, table2) %>%
  as.data.table()
```

The outcome of both functions is identical:

```{r}
a <- merge(table1, table2)
b <- inner_join(table1, table2) %>%
  as.data.table()
identical(a,b)
```

Let's check time of execution:

```{r}
rbenchmark::benchmark(merge(table1, table2),
                      inner_join(table1, table2) %>% as.data.table(),
                      inner_join(table1_tibble, table2_tibble),
                      order = "relative",
                      columns = c("test", "replications", "elapsed", "relative"))
```

We see that `dtplyr` is faster than `dplyr`, but slower than the original `data.table` code.

## Left join

The result of a *left join* is a table with all the rows of the left table (the first to be entered):

```{r}
merge(table1, table2, all.x = TRUE)
```

We can do the same with `left_join` in `dplyr` and `dtplyr`:

```{r}
left_join(table1, table2) %>%
  as.data.table()
```

## Rigth join

In a *right join* we retain all the rows of the right table:

```{r}
merge(table1, table2, all.y = TRUE)
```

It's easy to guess that we need to use `right_join` here:

```{r}
right_join(table1, table2) %>%
  as.data.table()
```

## Outer or full join

The result of a *full* or *outer join* is a table containing all the rows of both tables:

```{r}
merge(table1, table2, all.x = TRUE, all.y = TRUE)
```

The `dtplyr` function for a full join is `full_join`:

```{r}
full_join(table1, table2) %>%
  as.data.table()
```

## Cartesian merge

These two tables had many repeated values in the `id` column, so the result of merging them has many rows: 

```{r}
table3
table4
```

In `data.table`, if the result of the merge has more rows than the sum of rows of both tables, we need to add `allow.cartesian = TRUE`.

```{r}
merge(table3, table4, allow.cartesian = TRUE)
```

In dtplyr `allow.cartesian = TRUE` is set by default:

```{r}
inner_join(table3, table4) %>%
  as.data.table()
```

## References

Vignette of joining two tables in `dplyr` <https://dplyr.tidyverse.org/reference/join.html>

`dtplyr` package intro: <https://www.tidyverse.org/blog/2019/11/dtplyr-1-0-0/>

A comparison of `dplyr` and `data.table`, including merges: <https://atrebas.github.io/post/2019-03-03-datatable-dplyr/>

*I have used version 1.13.6 of data.table, version 1.0.5 of dplyr and 1.1.0 of dtplyr in R 4.0.3.*
