---
title: Ploting an Horizontal Bar Chart
author: Jose M Sallan
date: '2023-03-08'
slug: plotting-an-horizontal-bar-chart
categories:
  - R
tags:
  - data visualization
  - ggplot
  - R
meta_img: images/image.png
description: Description for the page
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

In this post, I will present the workflow to create an horizontal barchart presenting a set of values. A good practice for those tables is to arrange bars in decreasing order of the value. We'll see that we can to that with the `fct_reorder` function of `forcats`, included in the tidyverse.

I will be using the txhousing dataset, included in tidyverse, so I don't need more than that:

```{r}
library(tidyverse)
```

Let's see which cities of Texas have the most expensive housing. For each city, I am computing a price variable, equal to the median of median prices from 2010 onwards: 

```{r}
txhousing |>
  filter(year >= 2010) |>
  group_by(city) |>
  summarise(price = median(median, na.rm = TRUE)) |>
  arrange(-price)
```

Let's see the default plot of those values with `geom_bar(stat = "identity")`:

```{r}
txhousing |>
  filter(year >= 2010) |>
  group_by(city) |>
  summarise(price = median(median, na.rm = TRUE)) |>
  ggplot(aes(city, price)) +
  geom_bar(stat = "identity")
```

This plot is not nice, for several reasons:

* We cannot see the city names in the x axis.
* Bars are not arranged, so it is hard to see what are the most expensive cities.
* There are too many bars to see, which add little information if we focus on the more expensive cities.
* The standard output of ggplot has a lot of clutter.

We can get to see city names reversing axis. That's why we present an horizontal bar chart:

```{r}
txhousing |>
  filter(year >= 2010) |>
  group_by(city) |>
  summarise(price = median(median, na.rm = TRUE)) |>
  ggplot(aes(price, city)) +
  geom_bar(stat = "identity")
```

To reorder the cities, we use fct_reorder to change the city factor variable, so it is reordered by price:

```{r}
txhousing |>
  filter(year >= 2010) |>
  group_by(city) |>
  summarise(price = median(median, na.rm = TRUE)) |>
  mutate(city = fct_reorder(city, price)) |>
  ggplot(aes(price, city)) +
  geom_bar(stat = "identity")
```

If we want to pick the ten largest cities instead of all cities, we need to `arrange` the table by price, and then `slice` it to pick the first ten rows. Note that **`fct_reorder` reorders the chart, but not the table!**

```{r}
txhousing |>
  filter(year >= 2010) |>
  group_by(city) |>
  summarise(price = median(median, na.rm = TRUE)) |>
  arrange(-price) |>
  slice(1:10) |>
  mutate(city = fct_reorder(city, price)) |>
  ggplot(aes(price, city)) +
  geom_bar(stat = "identity")
```

Finally, we can improve the look of the table by:

* setting a blue color for bars with `fill` parameter in `geom_bar`.
* removing the backgroun and axis with `theme_minimal`.
* change the size of the title and axis text with `theme`.
* adding a descriptive enough title and removing axis labels with `labs`.

```{r}
txhousing |>
  filter(year >= 2010) |>
  group_by(city) |>
  summarise(price = median(median, na.rm = TRUE)) |>
  arrange(-price) |>
  slice(1:10) |>
  mutate(city = fct_reorder(city, price)) |>
  ggplot(aes(price, city)) +
  geom_bar(stat = "identity", fill = "#66B2FF") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 12),
        plot.title = element_text(size=15)) +
  labs(title = "The most expensive cities in Texas (median prices)", x = NULL, y = NULL)
```

The resulting chart is hopefully easier to read and to interpret than the default one.

```{r, echo=FALSE}
sessionInfo()
```

