---
title: Plotting a Multiple Line Plot
author: Jose M Sallan
date: '2023-03-15'
slug: plotting-a-multiple-line-plot
categories:
  - R
tags:
  - data visualization
  - ggplot
  - R
meta_img: images/image.png
description: Description for the page
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

It is frequent that we need to visualize the temporal evolution along time of one or several variables using a line plot. Doing multiple line plots with ggplot might not be easy at first, as usually we have each variable in a column. Here I will illustrate how to do that with the `economics` dataset, included in the tidyverse. I will also plot a variable and its rolling mean obtained with `zoo`.

```{r}
library(tidyverse)
library(zoo)
data("economics")
```

`economics` was produced from US economic time series data available from <https://fred.stlouisfed.org/>:

```{r}
economics
```

We see that we have multiple series of macroeconomic aggregates, presented on a monthly basis.

## A Single Line Plot

Let's start plotting the evolution of the personal savings rate `psavert`. This is straightforward with `geom_line()`.

```{r, fig.align='center'}
economics |>
  ggplot(aes(date, psavert)) +
  geom_line()
```

We can remove some clutter by applying `theme_minimal` and replacing axis labels with a descriptive title.

```{r, fig.align='center'}
economics |>
  ggplot(aes(date, psavert)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Temporal evolution of personal savings rate", x = NULL, y = NULL)
```

## Two or More Lines

Let's do a plot including personal savings rate `psavert` and median duration of unemployment `uempmed`. Each variable is presented in its column, so we need to transform data in two steps.

First, let's **`select` the variables included in the plot**:

* y-axis variables `psavert` and `uempmed` 
* the x-axis time variable `date`

```{r}
economics |>
  select(date, psavert, uempmed)
```

Second, **apply `pivot_longer` to have a long table** excluding the time variable.

```{r}
economics |>
  select(date, psavert, uempmed) |>
  pivot_longer(-date)
```

No matter how many variables had, now we have three columns: the x axis variable, `value` for the y axis and `name` to define the color of each line. Now we can do the plot.

```{r, fig.align='center'}
economics |>
  select(date, psavert, uempmed) |>
  pivot_longer(-date) |>
  ggplot(aes(date, value, color = name)) +
  geom_line()
```

Here is an improved version. I have defined line colors and legend labels with `scale_color_manual`, and placed the legend below the plot with `theme(legend.position = "bottom")`.

```{r, fig.align='center'}
economics |>
  select(date, psavert, uempmed) |>
  pivot_longer(-date) |>
  ggplot(aes(date, value, color = name)) +
  geom_line() +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("#FF8000", "#0080FF"), name = "variable", labels = c("savings", "unemployment")) +
  labs(title = "Temporal evolution of savings and unemployment", x= NULL ,y = NULL)
```

## Variable and Rolling Mean

A special case of a two-line plot is presenting a variable and its rolling mean. We can obtain that mean with `rollmean` from the `zoo` package.

```{r}
economics |>
  mutate(psavert_roll = rollmean(psavert, k = 12, fill = NA, align = "right")) |>
  select(date, psavert, psavert_roll) |>
  print(n = 15)
```

After that, let's **`select` the variables included in the plot**. Then, we use **apply `pivot_longer` to have a long table**.

```{r}
economics |>
  mutate(psavert_roll = rollmean(psavert, k = 12, fill = NA, align = "right")) |>
  select(date, psavert, psavert_roll) |>
  pivot_longer(-date) 
```

Now we are ready to do the plot:

```{r, fig.align='center'}
economics |>
  mutate(psavert_roll = rollmean(psavert, k = 12, fill = NA, align = "right")) |>
  select(date, psavert, psavert_roll) |>
  pivot_longer(-date) |>
  ggplot(aes(date, value, color = name)) +
  geom_line()
```

If we want to avoid the warning thrown by the `NA` of the rolling mean, we can remove these rows with `filter(!is.na(value))`. To plot the variable and its rolling mean, I have selected two colors with similar hue.

```{r, fig.align='center'}
economics |>
  mutate(psavert_roll = rollmean(psavert, k = 24, fill = NA, align = "right")) |>
  select(date, psavert, psavert_roll) |>
  pivot_longer(-date) |>
  filter(!is.na(value)) |>
  ggplot(aes(date, value, color = name)) +
  geom_line() +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("#99CCFF", "#0066CC"), name = "savings", labels = c("raw", "detrended")) +
  labs(title = "Temporal evolution of savings (raw and detrended)", x= NULL ,y = NULL)
```

Whenever you need to do a multi line plot in ggplot, do no forget the two steps:

* **`select` the variables included in the plot**
* **apply `pivot_longer` to have a long table**

```{r, echo=FALSE}
sessionInfo()
```
