---
title: Colors and shapes of points in ggplot2
author: Jose M Sallan
date: '2021-03-12'
slug: [colors-shapes-points]
categories:
  - R
tags:
  - ggplot
  - R
meta_img: images/image.png
description: Description for the page
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```


`ggplot` function returns plots with convenient default settings, at least for a first glance of results. If we want a prettier presentation, we can change those settings to suit our needs. Here we see an example of changing colors of an histogram: `fill` is the color of the bars, and `color` is the border of the bars:

```{r, message=TRUE, warning=TRUE}
library(ggplot2)
ggplot(iris, aes(Sepal.Length)) + 
  geom_histogram(bins=20, 
                 fill="blue", 
                 color="red")
```

This use of `fill` and `color` can be useful in some situations, but it is more frequent that we use colors and shapes to distinguish different levels of a categorical variable. Let's see some examples of this usage in scatterplots.

## Points with colors

Let's do a scatterplot of two variables of `iris`. I will use the color of the dots to distinguish observations of each species. To see the colors better, I have added a theme `theme_bw()` and increased the `size` of the points. I am doing by that specifying in the aesthetic mapping `aes` that the color of each observation is determined by its value of `Species`:

```{r}
ggplot(iris, aes(Sepal.Length, Sepal.Width, color=Species)) +
  geom_point(size=2) +
  theme_bw()
```

When a parameter like `color` is **inside the aesthetics**, we are not fixing what colors to use, but **what is the use of the color** in the plot (here, examining differences across the `Species` categorical variable).

To make the plot, `ggplot` uses a default palette of colors. If we are not happy with the result, we can change them with a **scale** (see a systematic presentation of scales at the [ggplot book](https://ggplot2-book.org/scales.html)).

Here I am using `scale_color_manual`, meaning that I will use a `manual` palette to define the colors for each value of `Species`. In the scale is also specified:

* the `name` of the legend,
* the `labels` to appear in the legend for each level of the categorical variable
* and the `values` of colors for each level of the categorical variable.

I have changed the three default values in this scale, using three different shades of blue to color the points.

```{r}
ggplot(iris, aes(Sepal.Length, Sepal.Width, color=Species)) +
  geom_point(size=2) +
  theme_bw() +
  scale_color_manual(name = "iris species", 
                     labels = c("iris setosa", "iris versicolor", "iris virginica"), 
                     values = c("#000099", "#0080FF", "#99CCFF"))
```

## Points with shapes

If your plot is in black and white, you might want to distinguish the categories with a shape. Shapes in points in R are defined with a number going from 0 to 25. You can learn about them in [this post](https://blog.albertkuo.me/post/point-shape-options-in-ggplot/) of Alberts Kuo's blog.

Here is the plot with the default values of shapes of points:

```{r}
ggplot(iris, aes(Sepal.Length, Sepal.Width, shape=Species)) +
  geom_point(size=2) +
  theme_bw()
```

We can change those default shapes with `scale_shape_manual`.

```{r}
ggplot(iris, aes(Sepal.Length, Sepal.Width, shape=Species)) +
  geom_point(size=2) +
  theme_bw() +
  scale_shape_manual(name = "iris species", 
                     labels = c("iris setosa", "iris versicolor", "iris virginica"), 
                     values = c(0, 3, 4))
```

We are seeing asterisks in some places because two observations of categories represented with [plus and cross](https://github.com/tidyverse/ggplot2/issues/2075) signs a are overlaid.

## Points with shapes and colors

We can assign a shape and a color to each category using a scale for each. If we use the same variable in both scales, we obtain a single legend. I have also enlarged point size to see the shapes better:

```{r}
ggplot(iris, aes(Sepal.Length, Sepal.Width, color=Species, shape=Species)) +
  geom_point(size=2) +
  theme_bw() +
  scale_shape_manual(name = "iris species", 
                     labels = c("iris setosa", "iris versicolor", "iris virginica"), 
                     values = c(0, 3, 4)) +
  scale_color_manual(name = "iris species", 
                     labels = c("iris setosa", "iris versicolor", "iris virginica"), 
                     values=c("#FF0000", "#009900", "#0000FF"))
```

*(For a listing of colors available through the [hex triplet](https://en.wikipedia.org/wiki/Web_colors#Hex_triplet) syntax, visit this [RapidTables](https://www.rapidtables.com/web/color/RGB_Color.html) page, or the [Web colors](https://en.wikipedia.org/wiki/Web_colors) Wikipedia page.)*

## Colors and shapes in ggplot

We can use colors and shapes to distinguish between elements of different values of a category introducing them in an aesthetic. We must keep in mind that we are **not specifying which colors to use, but defining the use of the color** in the plot If we are not satisfied with the default colors or shapes, we cqn add a scale to customize them. 

I have presented how to control the appearance of scatterplots built with `geom_point()` presenting different colors and shapes for the levels of a categorical variable. I have used `scale_color_manual` for colors, and `scale_shape_manual` for shapes.

Scales not only control the `values` of colors that we assign to each category, but also legend parameters like its `name` or title or the `labels` we assign to each category.