---
title: Introducing ggplot
author: Jose M Sallan
date: '2021-02-26'
slug: []
categories:
  - R
tags:
  - ggplot
  - R
meta_img: images/image.png
description: Description for the page
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

`ggplot2` is the package of the Tidyverse to produce elegant visualizations, based on the Grammar of Graphics. To generate a plot with ggplot2 we need:

* A **dataset** to pick data from. As with the rest of the Tidyverse, ggplot expects tidy data stored into a data frame.
* An **aestethic mapping** of the variables of the dataset. Here you will set the plot axis and how you will convey additional information (e.g., coloring points according to a categorical variable).
* One or several layers or geoms.

There are so much things you can do with ggplot, so we'll explore very basic possibilities here. Let's use the `iris` dataset:

```{r}
library(tibble)
iris_tibble <- tibble(iris)
iris_tibble
```

## A univariate plot

Let's examine the properties of the `Sepal.Width` variable with an histogram. To produce it, we need to:

* Specify the dataset, `iris_tibble`,
* specify the aesthetic mapping: here is a single variable `aes(Sepal.Width)`,
* define the histogram with `geom_histogram()`.

```{r}
library(ggplot2)
ggplot(iris_tibble, aes(Sepal.Width)) + 
  geom_histogram()
```

The default value of the number of bins is 30, which are too much bins for this variable. So we have set `bins=15`. Let's use `labs` to set the name of axis and give the plot a title:

```{r}
ggplot(iris_tibble, aes(Sepal.Width)) + 
  geom_histogram(bins=15) +
  labs(title="Distribution of sepal width", x="sepal width", y = "count")
```

## A bivariate plot

Let's now do a scatterplot to examine the relationship between sepal width and sepal length. Now the aesthetic will be `aes(Sepal.Length, Sepal.Width)` (ggplot is learning that there are continuous variables in both axis) and as we want a point for each observation we will use `geom_point()`. We'll also add title and axis with `lab`.

```{r}
ggplot(iris_tibble, aes(Sepal.Length, Sepal.Width)) + 
  geom_point() +
  labs(title = "Sepal length versus sepal width", x = "sepal length", y = "sepal width")
```

Let's try know to see if there are any differences between `Species`. To do so, we'll assign a different color to each point depending of its species. We can do that by defining:

* a specific aesthetic for the points: `geom_point(aes(color=Species))`,
* redefining the aesthetic for the whole plot: `ggplot(iris_tibble, aes(Sepal.Length, Sepal.Width, color=Species))`.

Note that here we are not specifying the actual colors to plot: we are assigning a different color to each observation according to its species. We'll see that `ggplot` will assign a default color to each species, and add a legend that allows us to distinguish the Species of each observation.

```{r}
ggplot(iris_tibble, aes(Sepal.Length, Sepal.Width, color=Species)) + 
  geom_point() +
  labs(title = "Sepal length versus sepal width", x = "sepal length", y = "sepal width")
```

Here we see that the setosa species is quite different from the other two species.

An alternative way of seeing this is by making a plot for each species using `facet_grid()`:

```{r}
ggplot(iris_tibble, aes(Sepal.Length, Sepal.Width)) + 
  geom_point() +
  facet_grid(. ~ Species) +
  labs(title = "Sepal length versus sepal width", x = "sepal length", y = "sepal width")
```

This has been a very basic introduction to what `ggplot` can do. In forthcoming ggplot-related posts we'll see:

* how to customize colors, shapes and legends with scales,
* how to define themes to customize the appearance of your plot,
* how to use other geoms.

## References

Wickham, H.; Grolemund, G. R for Data Science. <https://r4ds.had.co.nz/>

Wickham, H.; Navarro, D.; Pedersen, T. L. ggplot2: Elegant Graphics for Data Analysis. <https://ggplot2-book.org/>

Wilkinson, L. (2005). *The Grammar of Graphics.* Springer. Statistics and computing series. <https://www.springer.com/gp/book/9780387245447>