---
title: Analyzing a Lottery with Monte Carlo Simulation
author: Jose M Sallan
date: '2024-10-04'
slug: []
categories:
  - R
  - simulation
tags:
  - R
  - Monte Carlo
meta_img: images/image.png
description: Description for the page
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Monte Carlo simulation is a computational technique used to model the probability of different outcomes in a process that is inherently uncertain. It relies on repeated random sampling to obtain numerical results, simulating the effects of uncertainty and variability in mathematical models. The name of Monte Carlo was suggested by the physicist Nikolas Metropolis, while working at Los Alamos laboratory in the late 1940s, referring to Monte Carlo casino in Monaco.

The key steps in a Monte Carlo simulation are:

1. **Define the Problem**: Establish a mathematical model of the process or system you're studying.
2. **Identify Uncertainties**: Specify the variables in your model that are uncertain (e.g., future stock prices, interest rates, or demand for a product).
3. **Generate Random Inputs**: Use random or pseudo-random numbers to simulate the uncertain inputs. These inputs are typically drawn from probability distributions (e.g., normal, uniform, beta PERT).
4. **Run Simulations**: For each simulation run, input the random variables into the model and calculate the result. Repeat this process many times to cover a wide range of possible scenarios.
5. **Analyze Results**: Collect the outputs from all simulations to create a probability distribution of possible outcomes. Analyze this distribution to understand likely ranges, risks, and probabilities of different outcomes.

Here are some applications of the Monte Carlo simulation:

- **Finance**: Modeling the potential future returns of an investment portfolio, risk analysis, and option pricing.
- **Engineering**: Estimating system reliability or failure risks in complex systems.
- **Project Management**: Assessing the likelihood of completing a project within a certain time or budget.
- **Operations Research**: Optimizing processes under uncertainty, such as inventory management or production scheduling.

## A Simple Example

Let's illustrate the Monte Carlo method by examining a lottery presented by Kiehl Dang. Here is the **problem definition**:

*To win a certain lotto, a person must spell the word big. Sixty percent of the tickets contain the letter b, 30% contain the letter i, and 10% contain the letter g. Find the average number of tickets a person must buy to win the prize.*

We need to **identify uncertainties**. These are the values of the tickets, which follow the probability law expressed in the statement.

Then we need to **generate random inputs**. Let's write a R function to do this:

```{r}
run_big <- function(p1 = 0.6, p2 = 0.3){
  
  tickets <- runif(1)
  success <- FALSE
  
  while(!success){
    
    b <- any(tickets <= p1)
    i <- any(tickets > p1 & tickets <= p1 + p2)
    g <- any(tickets > p1 + p2)
    
    if(all(c(b, i, g))){
      success <- TRUE
    }else(
      tickets <- c(tickets, runif(1))
    )
  }
  
  return(length(tickets))
}
```

The function has two parameters: the probability of getting a ticket with letter **b** `p1`, and the probability of getting a ticket with letter **i** `p2`. The probability of getting a **g** is equal to `1 - p1 - p2`. The function builds a vector of `tickets` until it contains all three letters: when this happens it exits the loop making `success = TRUE`. The function uses `runif()` to generate a random number for each ticket. The function returns the length of `tickets`, which is the required number of tickets to obtain the three letters.

Once defined the function, we can **run the simulations** using the `sapply()` iterator. Here I am saving the values of the one thousand runs in the `trials` vector.

```{r}
set.seed(1111)
trials <- sapply(1:1000, \(x) run_big())
```

Now it is time to **analyze results**. Let's check the mean value of required tickets:

```{r}
mean(trials)
```

The obtained value is close to the obtained by Kiehl Dang with a large values of trials.

When performing a Monte Carlo simulation, we are not only interested in the mean value, but also in the probability distribution. Let's examine the values obtained from the simulation with an histogram.

```{r, fig.align='center'}
hist(trials)
```

While in most cases we get the three letters with a small number of tickets, sometimes we need to buy many of them to get the three letters: we have a **fat tailed** distribution. Let's use the `quantile()` function to evaluate risks.

```{r}
quantile(trials, seq(0, 1, 0.1))
```

Here we observe that although the mean value of `trials` is around eleven, there is a probability of 10% of needing to buy 21 tickets or more to obtain the three letters.

The `run_big()` function allows modifying the probability values of each letter. Intuitively, it seems that the more balanced the probabilities, the less tickets we will need to buy. We can check this running the function with `p1 = 1/3` and `p2 = 1/3`.

```{r}
set.seed(1313)
trials_balanced <- sapply(1:1000, \(x) run_big(p1 = 1/3, p2  = 1/3))
```

The mean number of tickets is now:

```{r}
mean(trials_balanced)
```

And the probability distribution:

```{r, fig.align='center'}
hist(trials_balanced)
```

We observe now that the average number of required tickets is much smaller than in the previous case, and that the extreme values are also smaller:

```{r}
quantile(trials_balanced, seq(0, 1, 0.1))
```

While in the previous case of unbalanced probabilities we have a small but relevant probability of needing to buy 21 tickets or more, this event is quite unlikely when probabilities are balanced. Therefore, we have confirmed our intuition.

## The Monte Carlo Simulation

Monte Carlo simulation is particularly valuable because it helps to account for the variability and randomness in real-world scenarios of processes too complex to be modeled exactly, providing a more robust understanding of potential risks and outcomes. When running a Monte Carlo simulation, we need to pay attention not only to central tendency statistics like mean or median, but also to the overall probability distribution.

## References

-   Dang, Kiehl (2022). *Monte Carlo simulation in R (Detailed code explanation)— Lottery Winner.* <https://bit.ly/MCLottery>
-   Monte Carlo method at Wikipedia. <https://en.wikipedia.org/wiki/Monte_Carlo_method>

The introduction to the Monte Carlo simulation method is an edition of a ChatGPT rendering.

## Session Info

```{r, echo=FALSE}
sessionInfo()
```


