---
title: Line and bar plots in ggplot2
author: Jose M Sallan
date: '2021-03-19'
slug: line_and_bar_plots_ggplot2
categories:
  - R
tags:
  - ggplot
meta_img: images/image.png
description: Description for the page
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```


It is frequent that we need to visualize relationships between variables with a **dependence relationship**, where the independent variable is in the horizontal axis, and the dependent variable in the vertical axis. Sometimes we are interested in the **absolute magnitude** of the dependent variable, and in other occasions in showing how the dependent variable **changes** as the independent variable changes. To plan our plots, we can take two recommendations from *Calling Bullshit* (Bergstronm and West, 2020). The first is about the type of chart to use:

> By its design, a bar graph emphasizes the absolute magnitude of values associated with each category, whereas a line graph emphasizes the change in the dependent variable (usually the y value) as the independent variable (usually the x value) changes. (Bergstronm and West, 2020:157).

The second is about vertical axis limits:

> While the bars in a bar chart should extend to zero, a line graph does not need to include zero in the dependent variable axis (Bergstronm and West, 2020:156-157).

In this post, we will see how can plot line and bar charts in `ggplot`, and how to include a categorical variable in those charts. To do so, I will use a time series of Spanish population data taken from the [Instituto Nacional de Estadística](https://en.wikipedia.org/wiki/Instituto_Nacional_de_Estad%C3%ADstica_(Spain)) (INE).


## Line charts to emphasize changes in the dependent variable

Let's pick a series of population data taken from the Spanish Labor Force Survey:

```{r}
library(tidyverse) #to load ggplot and dplyr
library(ESdata)    #to get the data
pop_quarterly <- epa_edad %>% 
  filter(region=="ES", edad=="total", sexo=="total", dato=="pob")
pop_quarterly
```

`pop_quarterly` is a data frame with `r nrow(pop_quarterly)` observations.  e will plot the independent variable `periodo` (in `Date` format) in the horizontal axis, and the dependent variable `valor` in the vertical axis, so that the aesthetic is `aes(periodo, valor)`.

We use `geom_line()` to plot line charts to emphasize **changes** of the dependent variable as the independent variable changes. `ggplot` is building by default a vertical axis that does not start from zero, following Bergstrom and West's recommendations. I have changed axis names and presented a title with `labs`, and changed the default theme to `theme_bw()`.

```{r}
pop_quarterly %>%
  ggplot(aes(periodo, valor)) +
  geom_line() +
  labs(title="Evolution of Spanish population", x="", y="population (thousands)") +
  theme_bw()
```

We observe Spanish population increased steadily from 2004 to approximately 2012, and that has started recovering by 2017 but with a slower pace than the previous growth cycle. Vertical axis ranges between 41 and 46 million people, which is the variation of population for this temporal series.

We are now interested in analyzing the evolution of Spanish population by gender. We need a dataset including the value of male and female Spanish population for each observation:

```{r}
pop_quarterly_gender <- epa_edad %>% 
  filter(region=="ES", edad=="total", sexo!="total", dato=="pob")
pop_quarterly_gender
```

The pop_quarterly_gender data frame has `r nrow(pop_quarterly_gender)`, twice the rows of the previous table. This is a dataset in **long format**, and it is [tidy data](https://jmsallan.netlify.app/blog/tidy_data/) because each row represents an observation. Remember that `tidyverse` functions expect tidy data to work.

By adding a `color` parameter to the aesthetic, `ggplot` will plot a line for each category. The value of `color` should be the variable defining the level of the categorical variable for each observation, which here is `sexo`. Now the aesthetic is `aes(periodo, valor, color=sexo)`.

By introducing the `color`in the aesthetic, we are not defining the colors, but defining what is the use of the color in the plot. To override `ggplot` defaults I have used `scale_color_manual` to change the legend title, the value of labels for each level of the categorical variable and the color of each line. I have used `theme(legend.position = "bottom")` to put the legend at the bottom of the plot, so that the chart can be seen larger than in the default legend position at the right.

```{r}
pop_quarterly_gender %>%
  ggplot(aes(periodo, valor, color=sexo)) +
  geom_line() +
  labs(title="Evolution of Spanish population by gender", x="", y="population (thousands)") +
  scale_color_manual(name="gender", labels=c("men", "women"), values =c("#0066CC", "#CC0000")) +
  theme_bw() +
  theme(legend.position = "bottom")
```

In this line chart we observe that there are consistently more women than men living in Spain, and that the decrease of population has been slightly larger for men than for women.

## Bar charts to emphasize absolute magnitude of the dependent variable

Let's examine now the evolution of **absolute** values of Spanish population. Instead of quarterly data, we will use a yearly dataset including the last observation of each year:

```{r}
library(lubridate) #to get the year of each observation
pop_yearly <- epa_edad %>% 
  filter(region=="ES", edad=="total", sexo=="total", dato=="pob") %>%
  mutate(year=year(periodo)) %>%
  group_by(year) %>%
  summarise(pop = last(valor), .groups = "drop")
pop_yearly
```

To examinethe absolute magnitude of the dependent variable, we build a bar plot using `geom_col()`. I have used `labs` and `theme_bw()` like in the previous plots and changed the color of the bars for aesthetic reasons with the `fill` parameter.

```{r}
pop_yearly %>% 
  ggplot(aes(year, pop)) +
  geom_col(fill="#808080") +
  labs(title="Evolution of Spanish population", x="", y="population (thousands)") +
  theme_bw()
```

With this bar plot, we focus on **absolute magnitudes** so we put changes in perspective. We realize that the variations of total population are relatively small respect to the total population.

Let's examine the yearly evolution of Spanish population by gender, building the adequate dataset in long format:

```{r}
pop_yearly_gender <- epa_edad %>% 
  filter(region=="ES", edad=="total", sexo!="total", dato=="pob") %>%
  mutate(year=year(periodo)) %>%
  group_by(year, sexo) %>%
  summarise(pop = last(valor), .groups = "drop")
pop_yearly_gender
```

To distinguish by gender, I have added `fill` to the aesthetic, and I have used `scale_fill_manual` to customize the plot. I have chosen now slightly paler shades of blue and red for aesthetic convenience.

```{r}
pop_yearly_gender %>%
  ggplot(aes(year, pop, fill=sexo)) +
  geom_col() +
  labs(title="Evolution of Spanish population by gender", x="", y="population (thousands)") +
  scale_fill_manual(name="gender", labels=c("men", "women"), values =c("#3399FF", "#FF3333")) +
  theme_bw() +
  theme(legend.position = "bottom")
```

In this barplot bars are **stacked**, meaning that we have a bar for each year with two colors, one for each category. This representation is adequate here, because the sum of both categories accounts for the total population. If we want to compare the values for each category in each year plotting bars for each category side by side, we can specify `position = "dodge"` inside `geom_col()`:

```{r}
pop_yearly_gender %>%
  ggplot(aes(year, pop, fill=sexo)) +
  geom_col(position = "dodge") +
  labs(title="Evolution of Spanish population by gender", x="", y="population (thousands)") +
  scale_fill_manual(name="gender", labels=c("men", "women"), values =c("#3399FF", "#FF3333")) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Area charts to emphasize the absolute magnitude of the depedent variable

To build the column plots, we have picked an observation for each year, as a bar plot with quarterly values has too many bars to visualize them correctly. An alternative to visualize absolute magnitude of a dependent variable with many observations is `geom_area()`. Let's examine the absolute magnitude of men and women population with an area plot:

```{r}
pop_quarterly_gender %>%
  ggplot(aes(periodo, valor, fill=sexo)) +
  geom_area() +
  labs(title="Evolution of Spanish population by gender", x="", y="population (thousands)") +
  scale_fill_manual(name="gender", labels=c("men", "women"), values =c("#3399FF", "#FF3333")) +
  theme_bw() +
  theme(legend.position = "bottom")
```

We have used `geom_area()` to examine the absolute magnitude of temporal evolution of population with `pop_quarterly_gender`, introducing `fill` in the aesthetic. 

## Examining relationships between dependent and independent variables

Whenever we are examining the relationship between two variables, the first thing we need to ask ourselves is if there is an evident dependence relatiosnship between them If there is not, it may be adequate to use a [scatterplot](https://jmsallan.netlify.app/blog/2021-03-12-colors-and-shapes-of-points-in-ggplot2/) to represent them. If the dependence relationship exists, the representation to choose will depend on our intentions:

* If we intend to examine **changes** of the dependent variable, we will use **line plots**. We can do that with `geom_line()` and examine different levels of a categorical variable including `color` in the aesthetic.
* If we want to examine the **absolute magnitude** of the dependent variable we will use **bar plots** with `geom_col()`, or **area plots** with `geom_area()`. In these plots, we use `fill` to examine the evolution of different levels of a categorical variable.

In both cases, `ggplot` will set by default the adequate range of values for the vertical axis. In line plots, vertical axis will be set to the range of values where change takes place. In bar and area plots, the vertical axis will start from zero.


## Reference

Bergstrm, Carl T. & West, Jevin D. (2020). *Calling bullshit: The art of skepticism in a data-driven world.* Random House, New York.

