---
title: Predicting with decision tress using rpart
author: Jose M Sallan
date: '2022-07-18'
slug: predicting-with-decision-tress-using-rpart
categories:
  - R
tags:
  - decision trees
  - machine learning
  - R
meta_img: images/image.png
description: Description for the page
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

In this post, I will make a short introduction to decision trees with the `rpart` package. This package implements the ideas about classification and regression trees presented in Breiman *et al*. (1983). I will present how `rpart` can be used for classification and numerical prediction, and how to plot the outcome of `rpart` using the `rpart.plot` package.

I will also use the `dplyr` and `ggplot2` for data manipulation and visualization, `BAdatasets` to access the `WineQuality` dataset, `mlbench` to access the `BostonHousing` dataset and `yardstick` to obtain classification metrics.


```{r}
library(dplyr)
library(ggplot2)
library(rpart)
library(rpart.plot)
library(BAdatasets)
library(mlbench)
library(yardstick)
```

Let's start with the same set of synthetic data we used for `C50`:

```{r}
n <- 400
set.seed(2020)
x <- c(runif(n*0.5, 0, 10), runif(n*0.25, 4, 10), runif(n*0.25, 0, 6))
y <- c(runif(n*0.5, 0, 5), runif(n*0.25, 4, 10), runif(n*0.25, 4, 10))
class <- as.factor(c(rep("a", n*0.5), c(rep("b", n*0.25)), c(rep("c", n*0.25))))

s_data <- tibble(x=x, y=y, class = class)
```

The `rpart` function uses the formula and data syntax. We can use `rpart.plot` to see the result:

```{r, fig.align='center'}
dt01 <- rpart(class ~  ., s_data)
rpart.plot(dt01)
```

This classification is very precise and complex, but also prone to overfitting. We can control how precise and therefore prone to overfitting a `rpart` tree is with the **complexity parameter** `cp`.  Any split that does not decrease the overall lack of fit by a factor of `cp` is not attempted. Small values of `cp` will create large trees and large values may not produce any tree. The default value of `cp` is of 0.01. Let's see the result to increase `cp` by an order of magnitude:

```{r, fig.align='center'}
dt02 <- rpart(class ~  ., s_data, cp = 0.1)
rpart.plot(dt02)
```

The outcome is similar to the obtained with `C50` and can represent a good balance between bias and variance. Other ways of controlling tree size can be found at `rpart.control`.

## Classifying with rpart

Let's build the `wines` table of features predicting quality of red and white Portugese wines from data of `WineQuality`:

```{r}
data("WineQuality")

red <- WineQuality$red %>%
  mutate(type = "red")
white <- WineQuality$white %>%
  mutate(type = "white")

wines <- bind_rows(red, white)
wines <- wines %>%
  mutate(quality = case_when(quality == 3 ~ 4,
                             quality == 9 ~ 8,
                             !quality %in% c(3,9) ~ quality))
wines <- wines %>%
  mutate(quality = factor(quality))
```

Let's see the tree we obtain with `rpart` defaults:

```{r, fig.align='center'}
dt_wines01 <- rpart(quality ~ ., wines)
rpart.plot(dt_wines01)
```

This classification looks too simplistic, as it has not been able to predict extreme values of quality. Let's try a much smaller value of `cp`:

```{r}
dt_wines02 <- rpart(quality ~ ., wines, cp = 0.0001)
```

The resulting tree is too large to be plot. Let's use the `predict` function with `type = "class"` to obtain the predicted values for each element of the sample.

```{r}
table_wines <- data.frame(value = wines$quality, pred = predict(dt_wines02, wines, type = "class"))
```

Here is the confusion matrix:

```{r}
conf_mat(table_wines, truth = value, estimate = pred)
```

And here some classification metrics:

```{r}
class_metrics <- metric_set(accuracy, precision, recall)
class_metrics(table_wines, truth = value, estimate = pred)
```

The classification metrics are not as good as with `C50`, as rpart does not implement winnowing and boosting. But maybe this classification is less prone to overfitting.

`rpart` offers a measure of **variable importance** equal to the sum of the goodness of split measures for each split for which it was the primary variable:

```{r}
dt_wines02$variable.importance
```

The variables used in early splits tend to have higher variable importance.

## Prediction of house prices in Boston Housing with rpart

Let's see now how can we use `rpart` for numerical prediction. I will be using the `BostonHousing` dataset presented by Harrison and Rubinfeld (1978). The target variable is `medv`, the median value of owner-occupied homes for each Boston census tract.
 
```{r}
data("BostonHousing")
BostonHousing %>%
  glimpse()
```

The syntax for numerical prediction is similar to classification: `rpart` knows the type of problem by the class of target variable. The predicted value for the observations in a leaf is the average value of the target variable for those observations.

```{r, fig.align='center'}
dt_bh <- rpart(medv ~ ., BostonHousing)
rpart.plot(dt_bh)
```

I have used the default `cp = 0.01`, obtaining a tree with a reasonable fit. Let's examine variable importance:

```{r}
dt_bh$variable.importance
```

To use the `yardstick` capabilities for numerical prediction, I am storing the real and predicted values of the target variable in a data frame.

```{r}
table_bh <- data.frame(value = BostonHousing$medv, 
                       pred = predict(dt_bh, BostonHousing))
```

Here I am obtaining the metrics, which show a reasonable fit.

```{r}
np_metrics <- metric_set(rsq, rmse, mae)
np_metrics(table_bh, truth = value, estimate = pred)
```

Here is the plot of real versus predicted values. The algorithm provides as many predicted values as terminal nodes of the tree.

```{r, fig.align='center'}
ggplot(table_bh, aes(value, pred)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  theme_minimal() +
  labs(x = "real values", y = "predicted values", title = "Prediction of Boston Housing with rpart")
```

The `rpart` package implements in R the decision tree techniques presented in the CART book by Breiman *et al.* (1983). The package can be used for classification and numerical prediction. An effective implementation calls for tuning of some of the parameters of `rpart.control`, for instance the complexity parameter `cp`.

## References

* Atkinson, E. J. The `rpart` package. <https://github.com/bethatkinson/rpart>
* Breiman, L., Friedman, J. H., Olshen, R. A., & Stone, C. J. (1983). *Classification and regression trees.* Wadsworth, Belmont, CA.
* Harrison, D. and Rubinfeld, D.L. (1978). Hedonic prices and the demand for clean air. *Journal of Environmental Economics and Management*, 5, 81–102.
* Milborrow, S. (2021). *Plotting `rpart` trees with the `rpart.plot` package.* <http://www.milbo.org/rpart-plot/prp.pdf>
* Therneau, T. M.; Atkinson, E. J. (2022). *An Introduction to Recursive Partitioning Using the RPART Routines.* <https://cran.r-project.org/web/packages/rpart/vignettes/longintro.pdf> Mayo Foundation.

## Session info

```{r, echo=FALSE}
sessionInfo()
```
