---
title: Solving linear programming with GLPK in R
author: Jose M Sallan
date: '2021-12-12'
slug: solving-linear-programming-with-glpk-in-r
categories:
  - R
tags:
  - optimization
  - linear programming
meta_img: images/image.png
description: Description for the page
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Rglpk)
```

Linear programming consist of optimizing a linear function subject to a set of linear constraints. If all or some of the decision variables are integer, it is called integer and mixed integer linear programming, respectively. Linear programming is a powerful optimization technique, and the first to consider when tackling combinatorial problems.

In this post, I will show how to solve linear programming in R using the `Rglpk` package. This is a R API for **GLPK**, the **G**NU **L**inear **P**rogramming **K**it. It is an open-source  set of routines written in ANSI C and organized in the form of a callable library.

Let's see how can we solve a linear programming formulation written in CPLEX or MathProg formats. We need two steps for doing this:

* Reading the model from a text file with `Rglpk_read_file`.
* Solving the model with `Rglpk_solve_LP`. The inputs of `Rglpk_solve_LP` are the outputs of `Rglpk_read_file`.

To make the task easier, I have written a `solve_lp` function that does the following:

* Reads the text file with path `filename`,
* Reads the text file of `type` `MathProg` or `CPLEX_LP`, and saves the result in the object `lp_model`.
* Obtains the solution `lp_solution` from the elements of `lp_model`.
* If the solution is feasible, returns the solution and `success = TRUE`. If not, returns `success = FALSE`. The solution is a list including a named vector `sol`, the value of the objective function `obj` and the `success` variable.

```{r}
solve_lp <- function(filename, type = "MathProg"){
  
  lp_model <- Rglpk_read_file(filename, type = type)

  lp_solution <- Rglpk_solve_LP(obj = lp_model$objective, 
                              mat = lp_model$constraints[[1]],
                              dir = lp_model$constraints[[2]],
                              rhs = lp_model$constraints[[3]],
                              types = lp_model$types,
                              bounds = lp_model$bounds,
                              max = lp_model$maximum)
  
  if(lp_solution$status == 0){
    sol <- lp_solution$solution
    names(sol) <- attr(lp_model, "objective_vars_names")
    obj = lp_solution$optimum
    return(list(sol = sol, obj = obj, success = TRUE))
  }else
    return(success = FALSE)
  
}
```

## A model in CPLEX format

Let's define a linear programming `model01` in CPLEX format. The bounds are not really necessary here, but are set to illustrate how to enter them in the model.

```{r}
model01 <- 
'Maximize
  cost: 3x1 + 1x2 + 3x3
Subject To
  r1: - x1 + 2x2 + x3 <= 4
  r2: 4x2 - 3x3 <= 2
  r3: x1 - 3x2 + 2x3 <= 3
Bounds
  0 <= x1 <= 99999
  0 <= x2 <= 99999
Integer
  x1 
Binaries
  x3
End'
```

Let's obtain the solution using `solve_lp`. Here I am creating the `model.mod` file with `writeLines` and deleting it after use with `unlink`.

```{r}
writeLines(model01, "model.mod")
solve_lp("model.mod", type = "CPLEX_LP")
unlink("model.mod")
```

## A model in MathProg format

Let's write the same model in `MathProg` format in `model01b`. Variable types and bounds are specified when defining variables.

```{r}
model01b <- 
'var x1, >=0, <=99999, integer;
var x2, >=0, <=99999;
var x3, binary;
maximize obj: 3 * x1 + x2 + 3 * x3;
s.t. r1: - x1 + 2 * x2 + x3 <= 4;
s.t. r2: 4 * x2 - 3 * x3 <= 2;
s.t. r3: x1 - 3 * x2 + 2 * x3 <= 3;
end;'
```

The solution of this model is:

```{r}
writeLines(model01b, "model.mod")
solve_lp("model.mod", type = "MathProg")
unlink("model.mod")
```

## A formulation in MathProg

Mathprog allows writing not only linear programs in a similar way as CPLEX, but also formulations of a generic problem that can be applied to different instances. Let's see the (abridged) formulation of the transportation problem:


```{r}
transport <- 'set I;
/* canning plants */

set J;
/* markets */

param a{i in I};
/* capacity of plant i in cases */

param b{j in J};
/* demand at market j in cases */

param d{i in I, j in J};
/* distance in thousands of miles */

param f;
/* freight in dollars per case per thousand miles */

param c{i in I, j in J} := f * d[i,j] / 1000;
/* transport cost in thousands of dollars per case */

var x{i in I, j in J} >= 0;
/* shipment quantities in cases */

minimize cost: sum{i in I, j in J} c[i,j] * x[i,j];
/* total transportation costs in thousands of dollars */

s.t. supply{i in I}: sum{j in J} x[i,j] <= a[i];
/* observe supply limit at plant i */

s.t. demand{j in J}: sum{i in I} x[i,j] >= b[j];
/* satisfy demand at market j */

data;

set I := Seattle San-Diego;

set J := New-York Chicago Topeka;

param a := Seattle     350
           San-Diego   600;

param b := New-York    325
           Chicago     300
           Topeka      275;

param d :              New-York   Chicago   Topeka :=
           Seattle     2.5        1.7       1.8
           San-Diego   2.5        1.8       1.4  ;

param f := 90;

end;'
```

The solution for this model is:

```{r}
writeLines(transport, "model.mod")
solve_lp("model.mod", type = "MathProg")
unlink("model.mod")
```

## Entering model parameters from the R environment

`Rglpk` can be seen as application programming interface (PI) to use GLPK through R. This means that the linear programming is solved using the C libraries of GLPK, returning the solution to the R environment. Instead of a CPLEX or Mathprog input, we can define the inputs for the API as R objects directly. 

Let's do that with the transportation problem of the previous section. I will define a `tp_lp` function for solving the transportation problem entering data from a named list with elements:

* `supply` or maximum capacity at each origin.
* `demand` from each destination.
* a `costs` matrix with cost coefficients.

Let's store the parameters of the model built in MathProg format in a named list with this structure:

```{r}
tp_sample <- list(supply = c(350, 600),
                  demand = c(325, 300, 275),
                  costs = matrix(c(2.5, 1.7, 1.8, 2.5, 1.8, 1.4)*90/1000, nrow = 2, ncol = 3, byrow = TRUE))
```

The function for solving the transportation problem with this input is:

```{r}
lp_tp <- function(tp){
  
  m <- length(tp$supply)
  n <- length(tp$demand)
  
  var_names <- character(n*m)
  
  for(i in 1:m)
    for(j in 1:n)
      var_names[(i-1)*n + j] <- paste0("x_", i, "_", j)
  
  obj <- c(t(tp$costs))
  rhs <- c(tp$supply, tp$demand)
  dir <- c(rep("<=", m), rep(">=", n))
  const <- matrix(0, n+m, n*m)
  
  for(i in 1:m){
    for(j in 1:n){
      const[i, n*(i-1) + j] <- 1
      const[m+j, n*(i-1) + j] <- 1
    }
  }
  
  lp_solution <- Rglpk_solve_LP(obj = obj, 
                                mat = const,
                                dir = dir,
                                rhs = rhs,
                                max = FALSE)
  
  if(lp_solution$status == 0){
    sol <- lp_solution$solution
    names(sol) <- var_names
    obj = lp_solution$optimum
    return(list(sol = sol, obj = obj, success = TRUE))
  }else
    return(success = FALSE)
  
}
```

Let's apply the function to our data, to check if the results is the same as in the previous section:

```{r}
lp_tp(tp_sample)
```

The obtained solution is the same as the previous section, but now with generic variable names generated by `lp_tp`.

## Acknowledgements

Thanks to Oriol Lordan for the model.mod trick, and to Vicenc Fernandez for the first example.

## References

* GLPK Wikibook <https://en.wikibooks.org/wiki/GLPK>
* GLPK (GNU Linear Programming Kit) <https://www.gnu.org/software/glpk/>


## Session info

```{r, echo=FALSE}
sessionInfo()
```

