---
title: Prim's Algorithm for the Minimum Spanning Tree
author: Jose M Sallan
date: '2024-05-18'
slug: prim-s-algorithm-for-the-minimum-spanning-tree
categories:
  - optimization
  - R
tags:
  - ggraph
  - optimization
  - tidygraph
meta_img: images/image.png
description: Description for the page
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this post, I will present a function to implement in base R the Prim's algorithm to obtain the minimum spanning tree (MST). I will use `tidygraph` and `ggraph` to plot the MST of a small graph example.

```{r, message=FALSE}
library(tidygraph)
library(ggraph)
```

For an undirected and weighted graph $G\left(N, E \right)$, the MST is a subset of the edges $E$ that connects all the nodes $N$ in the graph and with the minimum possible total edge weight. The MST is a problem with practical implications (e.g., wiring a set of cities at minimum cost) and it is also a building block of other algorithms like the Christofides algorithm for the Travelling Salesman Problem.

In the Prim's algorithm, we cover a node with the MST at each step.

-   **Step 1.** We initialize the set $L$ of not covered nodes with all the nodes of the graph.
-   **Step 2.** We remove from the set $L$ an arbitrary node $n_1$.
-   **Step 3.** We add to the MST the edge $e$ of minimum weight among those with a node $i \notin L$ and another node $j \in L$.
-   **Step 4.** We remove the node $i$ from the set $L$.
-   **Step 5.** We return to Step 3 until the set $L$ is empty.

Prim's algorithm is a greedy algorithm, which constructs a solution iteratively choosing at each iteration the most appealing element. The MST can be solved optimally with the Prim algorithm.

## Building the Function

The `prim()` function takes as argument a `D` distance matrix and returns a data frame with rows containing the nodes of each edge. It starts defining the `edges` table containing the edges ordered by non increasing distance. At each iteration we obtain the positions of the candidate edges `cand_edges`, and `new_edge`, the element of cand_edges of minimum distance. The output of the algorithm is the data frame `sol_edges`.

```{r}
prim <- function(d){
  
  n <- dim(d)[1]
  
  # setting edges
  edges <- data.frame(org = rep(1:n, times = n), dst = rep(1:n, each = n), dist = c(d))
  edges <- edges[which(edges$org != edges$dst), ]
  n_edges <- nrow(edges)
  edges <- edges[order(edges$dist), ]
  
  # setting vertices
  vertices <- 1:n
  notL <- 1
  sol_edges <- data.frame()
  
  for(i in 1:(n-1)){
    
    # candidate edges
    cand_edges <- which(edges$org%in% notL & !edges$dst %in% notL)
    
    # new edge: first compatible
    new_edge <- edges[cand_edges[1], ]
    
    #constructing the solution
    sol_edges <- rbind(sol_edges, new_edge)
    
    #updating set L
    notL <- c(notL, new_edge$dst)
    
  }
  
  sol_edges <- sol_edges[, 1:2]
  return(sol_edges)
  
}
```

## Applying the Function

Let's define a graph of six nodes with coordinates:

```{r}
l <- matrix(c(1, 2,
              2, 1,
              3, 2,
              4, 4,
              2, 5,
              4, 6), ncol = 2, byrow = TRUE)
```

The distance between nodes is obtained as:

```{r}
d_l <- as.matrix(dist(l))
```

Then, we apply the algorithm to obtain the MST with the prim() function:

```{r}
p_l <- prim(d_l)
p_l
```

## Plotting the MST

We can use the output of the algorithm to define a graph. We will label the nodes with letters `A` of `F`.

```{r}
g_l <- tbl_graph(edges = p_l)
g_l <- g_l |>
  activate(nodes) |>
  mutate(name = LETTERS[1:nrow(l)])
g_l
```

And now we can plot the MST.

```{r, fig.align='center'}
ggraph(g_l, layout = l) +
  geom_node_text(aes(label = name), size = 7) +
  geom_edge_link(start_cap = circle(5, "mm"), 
                 end_cap = circle(5, "mm"),
                 linewidth = 2, color = "#A0A0A0") + 
  theme_graph()
```

## Session Info

```{r, echo=FALSE}
sessionInfo()
```


