---
title: Introducing the beta probability distribution
author: Jose M Sallan
date: '2022-10-03'
slug: introducing-the-beta-distribution
categories:
  - R
  - statistics
tags:
  - ggplot
  - R
  - distributions
meta_img: images/image.png
description: Description for the page
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

In this post, I will present the beta distribution of probability, and illustrate how to use the `tidiyverse` to represent probability distributions. Beta distribution functions are included in the R base, so I will only be using the `tidyverse`, complemented with the `patchwork` package.

```{r}
library(tidyverse)
library(patchwork)
```

In probability theory and statistics, the **beta distribution** is a family of continuous probability distributions defined on the interval $\left[0, 1\right]$. It is used to model distribution of the probability of a specific success. 

The shape of the beta distribution is defined with two parameters $\alpha \geq 0$ and $\beta \geq 0$. The formula of the beta probability distribution is:

\[ d = \frac{x^{\alpha-1}\left(1-x\right)^{\beta-1}}{B\left(\alpha, \beta\right)} \]

The $B\left(\alpha, \beta\right)$ is a fixed term to scale the area under the function to one, and it is defined using the gamma function:

\[B\left(\alpha, \beta\right) = \frac{\Gamma\left(\alpha\right)\Gamma\left(\beta\right)}{\Gamma\left(\alpha+\beta\right)}\]

With the tidyverse, we can use the `dbeta` function in stat_function to plot a probability distribution function. Parameters `shape1` and `shape2` of dbeta correspond with $\alpha$ and $\beta$.

```{r, fig.align='center'}
  ggplot() +
  stat_function(fun = dbeta, args = list(shape1 = 10, shape2 = 40)) +
  theme_minimal()
```


## Interpretation of the beta distribution

The beta distribution:

\[ \frac{1}{B\left(\alpha, \beta\right)} x^{\alpha-1}\left(1-x\right)^{\beta-1}\]

is similar to the binomial distribution:

\[{n\choose x}  p^{x}\left(1-p\right)^{n-x}\]

While the binomial gives the distribution of number of successes given a probability, the beta gives the distribution of probability of success given the number of successes and failures. Then, $\alpha - 1$ is the number of successes and $\beta - 1$ is the number of failures. This interpretation of beta is valid for integer values of $\alpha0$ and $\beta0$.

Let's represent some beta distributions to illustrate this interpretation. First I am creating the `table_beta` function, that draws a the beta distribution for a number of sucesses `s` and failures  `f`. The other parameter is a label `l`. I am using  To draw the probability distribution, I create the variable `x` as a sequence of 100 values between 0 and 1, and then apply the `dbeta` function to obtain the value of density of probability `d`.

```{r}
table_beta <- function(s, f, l){
  t <- tibble(x = seq(0, 1, length.out = 100),
              s = s,
              f = f,
              sample = l) %>%
    mutate(d = dbeta(x, shape1 = s + 1, shape2 = f + 1))
  return(t)
}
```

The `table_plot` function plots a set of beta distributions. The value of `s`, `f` and `l` for each distribution is stored in a table `t`. Each element of the set needs a different value of the label `l`. `l` is passed as `color` in the `aes` of the plot, so that we obtain one line for each value of `l`.

```{r}
table_plot <- function(t){
  
  l <- lapply(1:nrow(t), \(i) table_beta(t$s[i], t$f[i], t$l[i]))
  
  df <- bind_rows(l)
  
  ggplot(df, aes(x, d, color = sample)) +
    geom_line() +
    theme_minimal() +
    scale_color_brewer(palette = "Reds", direction = -1)
}
```

Let's generate two sets of distributions. `t1` is a set of distribution with 20% of successes with a growing number of observations. `t2` is similar, but with a 80% of successes.

```{r}
t1 <- tibble(s = c(10, 50, 100, 200),
             f = c(40, 200, 400, 800),
             l = factor(s + f))

t2 <- tibble(s = c(40, 200, 400, 800),
             f = c(10, 50, 100, 200),
             l = factor(s + f))
```

Here I am presenting the plots of both families of distributions, `t1` on the left and `t2` on the right. They are put together thanks to the `patchwork` package.

```{r, out.width='100%', fig.height = 4}
p1a <- table_plot(t = t1) + 
  geom_vline(xintercept = 0.2, color = "#004C99") +
  ggtitle("Samples with 20% of success")

p1b <- table_plot(t = t2) + 
  geom_vline(xintercept = 0.8, color = "#004C99") +
  ggtitle("Samples with 80% of success")

p1a + p1b
```

In both cases, the variability of the probability distribution decreases with the number of observations. This is an example of increase of statistical power with larger sample sizes. 

We can see the effect of sample size on power calculating ranges of probabilities using `pbeta`. When we have 10 successes and 40 failures, the probability that the probability of success ranges between 0.21 and 0.19 is:

```{r}
pbeta(0.21, 11, 41) - pbeta(0.19, 11, 41)
```

But when we have 1000 successes and 4000 failures, the probability is:

```{r}
pbeta(0.21, 1001, 4001) - pbeta(0.19, 1001, 4001)
```

For additional clarification, I will examine a set of beta distributions with one hundred observations, but with different proportions of success and failure:

```{r, fig.align='center'}
t3 <- tibble(s = c(10, 20, 50, 80, 90),
             f = c(90, 80, 50, 20, 10),
             l = as.factor(c(0.1, 0.2, 0.5, 0.8, 0.9)))

table_plot(t = t3) +
  ggtitle("Several samples of 100 observations")
```

As the number of successes increases, the density of probability distribution leans to the right.

## Beta distribution with alpha or beta smaller than one

The formula of the beta distribution also admits values of shape parameters $\alpha$ and $\beta$ between one and zero. To examine these distributions, I will create a new `table_beta2` function taking `alpha` and `beta` as arguments...

```{r}
table_beta2 <- function(alpha, beta, l){
  t <- tibble(x = seq(0, 1, length.out = 100),
              alpha = alpha,
              beta = beta,
              sample = l) %>%
    mutate(d = dbeta(x, shape1 = alpha, shape2 = beta))
  return(t)
}
```

.. and its corresponding `table_plot2` function:

```{r}
table_plot2 <- function(t){
  
  l <- lapply(1:nrow(t), \(i) table_beta2(t$alpha[i], t$beta[i], t$l[i]))
  
  df <- bind_rows(l)
  
  ggplot(df, aes(x, d, color = sample)) +
    geom_line() +
    theme_minimal() +
    scale_color_brewer(palette = "Reds", direction = -1)
}
```

Let's examine first a `t4` set with a combination of fractional `alpha` and integer `beta` and a symmetrical `t5` set.

```{r, out.width='100%'}
t4 <- tibble(alpha = seq(0.1, 0.8, 0.2),
             beta = 2,
             l = as.factor(alpha))

t5 <- tibble(alpha = 2,
             beta = seq(0.1, 0.8, 0.2),
             l = as.factor(beta))


table_plot2(t4) + table_plot2(t5)
```

Being $\alpha$ much smaller than $\beta$, the probability distribution is skewed to the left. The swap of values of  shape parameters would lead to a symmetrical distribution.

Let's examine now two set of $\alpha$ and $beta$ smaller than one. In the `t6` set (left plot), both shape parameters add one, while in the `t7` set (right plot) are equal. In both plots, labels are the values of $\alpha$.

```{r, out.width='100%', fig.height = 4}
t6 <- tibble(alpha = seq(0.1, 0.8, 0.2),
             beta = 1 - alpha,
             l = as.factor(alpha))

t7 <- tibble(alpha = seq(0.1, 0.8, 0.2),
             beta = alpha,
             l = as.factor(alpha))

table_plot2(t6) + table_plot2(t7)
```

We observe U-shaped distributions, meaning that the distribution is skewed to extreme values of probability. In the StackExchange question listed in the references, one respondent provides an interpretation of this distribution based on a extrapolation of Polya urns to fractional values.

## References

* Kim, Aerin (2020). *Beta Distribution — Intuition, Examples, and Derivation.* <https://towardsdatascience.com/beta-distribution-intuition-examples-and-derivation-cf00f4db57af>
* StackExchange. *Why does the beta distribution become U shaped when $\alpha$ and $\beta$ <1?* <https://math.stackexchange.com/questions/3494530/why-does-the-beta-distribution-become-u-shaped-when-alpha-and-beta-1>

## Session info

```{r, echo=FALSE}
sessionInfo()
```

