---
title: Data manipulation with dplyr
author: Jose M Sallan
date: '2021-02-19'
slug: [dplyr-intro]
categories:
  - R
tags:
  - data structures
  - dplyr
meta_img: images/image.png
description: Description for the page
---

`dplyr` is a grammar of data manipulation for efficient transformation of rectangular data stored in data frames or tibbles. `dplyr` is part of the [Tidyverse](https://www.tidyverse.org/), and its functions expect tidy data.

```{r, message=FALSE, warning=FALSE}
library(dplyr)
```

An example of a tidy dataset is `iris`. Here we have defined a tibble version called `iris_table` for clarity, but you can use `dplyr` functions into `iris` straight.

```{r, message=FALSE, warning=FALSE}
library(tibble)
iris_tibble <- tibble(iris)
iris_tibble
```

## The basic verbs for data manipulation

Basic `dplyr` functions are implemented as a set of verbs:

* add new variables (columns) with `mutate()`
* pick variables (columns) with `select()`
* pick cases (rows) based on their values with `filter()`
* obtain summary statistics of a varaible wiht `summmarise()`
* order the rows according with variable values with `arrange()`

Here are some examples of usage of basic `dplyr` verbs:

```{r}
# add a new variable
iris_tibble <- mutate(iris_tibble, new_var = Sepal.Length/Sepal.Width)

# select two columns
select(iris_tibble, Sepal.Length, Sepal.Width)

# filter observations with Sepal.Length greater than 5.5
filter(iris_tibble, Sepal.Length > 5.5)

# obtain the mean of Sepal.Width
summarise(iris_tibble, m = mean(Sepal.Width))

# order by decreasing value of Sepal.Width
arrange(iris_tibble, desc(Sepal.Width))
```

## Piping operator

We can combine several dplyr verbs in a single instruction using the piping operator `%>%`:

```{r}
# obtain the mean of Sepal.Width for observations with Sepal.Length greater than 5.5

iris_tibble %>% 
  filter(Sepal.Length > 5.5) %>% 
  summarise(m=mean(Sepal.Width))
```

## Grouping

Sometimes we want to examine the properties of a dataset for each of the levels of a categorical variable (or combinations of levels). We can do that with `group_by`. It is often useful combining grouping with summarising:

```{r}
# mean of variables for each species
iris_tibble %>% 
  group_by(Species) %>%
  summarise(m_sl = mean(Sepal.Length), m_sw = mean(Sepal.Width), m_pl = mean(Petal.Length), m_pw = mean(Petal.Width))
```

## More functions

`dplyr` has many other functions for data manipulation. You can find them in the [dplyr tidyverse website](https://dplyr.tidyverse.org/) or in the [dplyr cheatsheet](https://github.com/rstudio/cheatsheets/blob/master/data-transformation.pdf).

