---
title: Merging with data.table
author: Jose M Sallan
date: '2021-04-09'
slug: merging-with-data-table
categories:
  - R
tags:
  - R
  - data structures
  - data.table
meta_img: images/image.png
description: Description for the page
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

When performing tasks of exploratory data analysis, sometimes we need to join tables by common variables. This is an operation similar to the defined in [relational algebra](https://en.wikipedia.org/wiki/Relational_algebra), although joining variables may not represent a unique ID for each row. Here I will illustrate how to merge tables using the `data.table` package, that allows handling large datasets in R effectively.

```{r}
library(data.table)
```

```{r, echo=FALSE}
table1 <- data.table(id = c("a", "a", "b", "b", "c", "e"), 
                     v = 1:6,
                     date = as.IDate(c("2021-01-01", "2021-03-23", "2021-02-13", "2021-05-28", "2021-12-17", "2021-12-31")))

table2 <- data.table(id = c("a", "b", "c", "d"), w = 1:4)

table3 <- data.table(id = c("a", "a", "a", "b", "c", "f"), g = 21:26)

table4 <- data.table(id = c(rep("a", 5), rep("b", 3), rep("c")), h = 31:39)

table5 <- data.table(id = c(rep("a", 5), rep("b", 3), rep("c")), 
                     h = 31:39,
                     date = as.IDate(c("2021-01-01", "2021-02-04", "2021-03-23", "2021-06-03", "2021-08-15", "2021-02-13", "2021-05-28", "2021-07-12", "2021-12-17")))
```


Let's consider the following tables.


```{r}
table1
table2
```

`table1` and `table2` can be merged by the column `id`. The most generic function to perform joins in `data.table` is `merge`, similar to the R base function of the same name. Let's see how to perform different merges with these two tables.

### Inner join

The result of an *inner join* is a table with the rows with values of merging variables existing in both tables. This is the default mode of `merge`, with `all.x = FALSE` and `all.y = FALSE`.

```{r}
merge(table1, table2)
```

`merge` has chosen the variables to merge picking the ones that have the same name in both tables. In `data.table`, we can control which are the merging variables using `setkey`:

```{r}
table1_keyed <- copy(table1)
table2_keyed <- copy(table2)
setkey(table1_keyed, "id")
setkey(table2_keyed, "id")
merge(table1_keyed, table2_keyed)
```

Let's see if we gain speed setting the key variables:

```{r}
rbenchmark::benchmark(merge(table1, table2),
                      merge(table1_keyed, table2_keyed),
                      order = "relative",
                      columns = c("test", "replications", "elapsed", "relative"))
```

Seeing the results, we better set the keys of `table1` and `table2`.

```{r}
setkey(table1, "id")
setkey(table2, "id")
```


### Left join

The result of a *left join* is a table with all the rows of the first table (the one on the left). The columns of the second table (on the right) with no correspondence with the first have `NA` in the merged table. We can perform a left join with `data.table` using `merge` with `all.x = TRUE`.

```{r}
merge(table1, table2, all.x = TRUE)
```

## Right join

The *right join* is analogous to the left join, but the resulting table contains all the rows of the second table:

```{r}
merge(table1, table2, all.y = TRUE)
```

In `data.table`, we have an alternative syntax for right joins using the `on` operator:

```{r}
table1[table2, on = .(id)]
```

we can get rid of the `on` operator if we define the keys of each table with `setkey`:

```{r}
table1[table2]
```

## Outer or full join

The result of a *full* or *outer join* is a table containing all the rows of both tables:

```{r}
merge(table1, table2, all.x = TRUE, all.y = TRUE)
```

## Cartesian join

A relational algebra requires that the columns used to merge variables have an unique value for each row. This is not always the case when we merge tables: we have seen than `table1` has several rows with `id` equal to `a` or `b`. Let's see what happens when we perform an inner join with a table3 with repeated values in its `id` column:

```{r}
table1
table3
setkey(table3, "id")
merge(table1, table3)
```

`merge` returns a table with all combinations of variables in the `id` column. The resulting table has more rows than any of the two rows. This is something similar to the [cartesian join in SQL](https://www.tutorialspoint.com/sql/sql-cartesian-joins.htm), but here the cartesian product is only between rows of both tables with the same value in the merging columns.

Let's consider a `table4` to `merge` with `table3`.

```{r}
table4
setkey(table4, "id")
```

If we try to perform `merge(table3, table4)`, `data.table` returns an error, because the number of rows of the merged table is larger than the sum of the rows of merged tables: `table3` has 6 rows, `table4` has 9 rows and the merged table has 19 rows. This is not exactly an error, but a warning that the size of the merged table can be too large. We must explicitly set `allow.cartesian = TRUE` to tell `data.table` that we are performing a cartesian join.

```{r}
merge(table3, table4, allow.cartesian = TRUE)
```

We can use the on syntax for cartesian merges (we `table3[table4, on =.(id), allow.cartesian = TRUE]` if the merging variables are not defined with `setkey`):

```{r}
table3[table4, allow.cartesian = TRUE]
```

## Rolling join

`data.table` has an additional feature to merge tables when we merge tables with dates or times. To illustrate that, let's consider merging `table1` and a new `table5`:

```{r}
table1
setkey(table1, "id", "date")
table5
setkey(table5, "id", "date")
```

When we perform a merge with the `on` syntax, we have `NA` in the columns of `table1` (what we have performed is equivalent to `table3[table4, on =.(id), allow.cartesian = TRUE]`):

```{r}
table1[table5]
```

Both tables have a column of temporal values. If we interpret that `date` is the moment when the variable `v` changes in `table1`, then the value of `v` in the second row of the merged table is `1`, `2` in rows 4 and 5 and `4` in row 8. We can implement that doing `roll = TRUE`:

```{r}
table1[table5, on = .(id, date), roll = TRUE]
```
