---
title: Dependency networks of installed R packages
author: Jose M Sallan
date: '2022-05-16'
slug: dependency-networks-of-installed-r-packages
categories:
  - R
tags:
  - complex networks
  - data.table
  - dplyr
  - ggraph
  - tidygraph
  - R
meta_img: images/image.png
description: Description for the page
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

R packages are bundles of code, data, documentation and test structured to be easily shared through repositories like CRAN. An R package usually needs other packages to work, or maybe needs some other packages for optional functionalities. There are three ways of specifying these dependencies in the `DESCRIPTION` file of a package:

* Prior to the roll-out of namespaces in R 2.14.0 in 2011, `Depends` was the only way to establish that the package *depends* on another package. Nowadays, dependencies are listed in `Imports`. We can use `Depends` to state a minimum version for R itself, e.g. `Depends: R (>= 3.6.0)`.
* Packages listed in `Imports` are needed by package users at runtime. Any time the package is installed, those packages will also be installed, if not already present.
* Packages listed in `Suggests` are either needed for development tasks or might unlock optional functionalities. They are not automatically installed along with the package.

The relationship dependencies between a set of R packages leads to the definition of a directed network of dependencies, where nodes are packages connected through a direct link *(i,j)* if package *j* has some dependency on package *i*.

To explore how these networks of dependencies look like, I will examine the network of packages installed in my computer. I will use `tidygraph` and `ggraph` for graph manipulation and plotting, and `data.table` and `dplyr` for tabular data handling.

```{r}
library(data.table)
library(dplyr)
library(tidygraph)
library(ggraph)
library(patchwork)
library(kableExtra)
```

I have checked the packages installed in my computer using `installed.packages()`. The output is a matrix, that I transform to a `data.table` object. This list of packages will be different for every R user, and will be representative of its preferences and of how he or she is using R.

```{r}
ip <- data.table(installed.packages())
```

Dependencies are listed in the `Depends`, `Imports` and `Suggests` columns. Let's see the `Depends` column:

```{r}
ip[, .(Package, Depends)]
```

I am building a table of dependencies with the `get_table_packages()` function:

* Selects a specific `column` of dependencies. 
* gets each of the packages of each row in the column with `data:table::tstrsplit`.
* Transforms the values of each package so that it represents a package name only. For instance, `dplyr (>= 0.8.5)` is transformed into `dplyr` and `R (>= 2.10.0)` into `R`.
* Pivots the table into a long table with `origin` and `destination` columns.
* Adds a `relation` column to store all data in a single table.

```{r}
get_table_packages <- function(dt, column){
  
  table <- copy(dt)
  vars <- c("Package", column)
  
  table <- table[, ..vars]
  
  table <- cbind(table[, 1], table[ , tstrsplit(table[[2]], ",")])
  table <- table[, lapply(.SD, \(x) gsub("^ ", "", x))]
  table <- table[, lapply(.SD, \(x) gsub(">=", " ", x))]
  table <- table[, lapply(.SD, \(x) gsub("\n", "", x))]
  table <- table[, lapply(.SD, \(x) gsub("\\(", "", x))]
  table <- table[, lapply(.SD, \(x) sapply(strsplit(x, " "), \(x) x[1]))]
  table <- melt(table, id.vars = "Package", na.rm = TRUE)
  
  table[, variable := NULL]
  table[, relation := column]
  setnames(table, c("destination", "origin", "relation"))
  setcolorder(table, c("origin", "destination", "relation"))
  
  return(table)
  
}
```

Finally I am applying the function to each dependency and storing the results in `ip_table`.

```{r}
rel_packages <- c("Depends", "Imports", "Suggests")

ip_list <- lapply(rel_packages, \(x) get_table_packages(ip, x))

ip_table <- rbindlist(ip_list)

rm(ip_list)
```

## Obtaining package networks

Let's obtain the three dependency networks defined in `rel_packages` and store then in a list. I am also calculating three node measures:

* in-degree `d_in`, the number of edges incident to a node.
* out-degree `d_out`, the number of edges going out of a node.
* betweenness `btw`, the number of shortest paths passing through a node.

```{r}
network_packages <- lapply(rel_packages, function(x){
  
  g <- tbl_graph(edges = ip_table[relation == x], directed = TRUE)
  
  g <- g %>%
    activate(nodes) %>%
    mutate(d_in = centrality_degree(mode = "in"),
         d_out = centrality_degree(mode = "out"),
         btw = centrality_betweenness())
    
  return(g)
})

names(network_packages) <- rel_packages
```

Let's examine each of the produced networks.

```{r}
network_packages
```

We observe that:

* The networks of `Depends` and `Imports` are trees. This makes sense since there may not be cyclical dependencies or imports. The set of nodes of both networks is a subset of the installed packages.
* The network of `Suggests` has cycles. The set of nodes is larger than the one of installed packages, indicating that some suggested packages are not installed.

## Plotting the networks

I am defining a `plot_network_packages` function to plot the networks. Note how I am choosing a very low value of transparency `alpha` for edges, as networks are relatively dense.

```{r}
plot_network_packages <- function(i){
  ggraph(network_packages[[i]], layout = "sugiyama") +
  geom_node_point(aes(label = name)) +
  geom_edge_link(alpha = 0.1, start_cap = circle(3, 'mm'), end_cap = circle(3, 'mm'), arrow = arrow(length = unit(2, 'mm'))) +
  theme_graph() +
  labs(title = paste("Network of", names(network_packages)[i]))
}

network_plots <- lapply(1:3, plot_network_packages)
```

To plot all networks at once, I am using `wrap_plots` from the `patchwork` package. Functions from `gridExtra` do not seem to work well with `ggraph` outcomes.

```{r, out.width='100%', fig.height = 12}
wrap_plots(network_plots, ncol = 1)
```

We observe that `Depends` and `Imports` have a tree like structure, more complex for `Imports`. Although the `Suggests` network has cycles and disconnected components, seems to behave like a tree for most of its nodes.

## Relevant nodes

Seeing that the networks of packages have a tree-like structure, we can establish two criteria to select relevant nodes:

* **Root nodes:** the packages at the top of the tree seem to be critical for the functionality of the system. Root nodes will have **in-degree equal to zero**, and high values of **out degree**.
* **Intermediate nodes:** for paths of relationships of two or more edges, the nodes in the middle are also important for package functionality. These nodes will have high values of **betweenness**.

Let's create a function to present the relevant packages of a network.

```{r}
get_table_measure <- function(i, nnodes = 5){
  
  node_table <- network_packages[[i]] %>%
    activate(nodes) %>%
    as_tibble()
  
  root <- node_table %>%
    filter(d_in == 0) %>%
    arrange(-d_out) %>%
    mutate(relation = names(network_packages)[i]) %>%
    select(relation, name, d_out) %>%
    rename(root = name) %>%
    slice(1:nnodes)
  
  interm <- node_table %>%
    arrange(-btw) %>%
    select(name, btw) %>%
    rename(intermediate = name) %>%
    slice(1:nnodes)
  
  table <- bind_cols(root, interm)
  
  return(table)
  
}
```

Here is the result of applying the function, presented in a table formatted with `kableExtra`.

```{r, fig.align='center'}
nodes_list <- lapply(1:3, \(i) get_table_measure(i))
nodes_table <- bind_rows(nodes_list)

nodes_table %>%
  kbl() %>%
  kable_paper(full_width = FALSE) %>%
  row_spec(1:5, background = "#FFFFCC") %>%
   row_spec(6:10, background = "#CCFFFF") %>%
   row_spec(11:15, background = "#FFCCCC")
  
```

Root and intermediate packages are different for each network, representative of each relationship. The main root package in `Depends` is the minimal version of R required. This was to be expected, given the role of `Depends` on package dependencies definition. Root packages in `Imports` are related with the tidyverse like magrittr or with data visualization like `grDevices`. Root packages in Suggests are related with package development, and most of them are not installed in my computer.

As for intermediate packages, the ones in `Depends` are not quite representative, as this relationship is not intended to be chained at several levels. The results of `Imports` are more informative, and show the relevance of the tidyverse in package development, at least the ones in my computer. This is also evident in `Suggests`, where together with package development appear other packages related with publishing and visualization like `knitr` and `ggplot2`.

The results of this analysis are not representative of the whole R CRAN package ecosystem. They have to be considered as a preliminary analysis for a further examination of the whole CRAN network.

## References

* *Depends or imports?* (From the *Developing R packages* course)
 <https://campus.datacamp.com/courses/developing-r-packages/checking-and-building-r-packages?ex=8>
* *Dependencies: What does your package need?* (from Wickham, H. & Bryant, J. *R packages*) <https://r-pkgs.org/description.html#description-dependencies>

## Session info

```{r, echo=FALSE}
sessionInfo()
```

