---
title: A Faceted Horizontal Bar Chart
author: Jose M Sallan
date: '2024-09-05'
slug: a-faceted-horizontal-bar-chart
categories:
  - R
tags:
  - data visualization
  - ggplot
  - purrr
  - R
meta_img: images/image.png
description: Description for the page
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

In this post, I will present how to plot a faceted bar plot. The challenge in defining such a graph is that the ordering of elements across facets can change. To allow this, we need the `reorder_within()` and "scale_y_reordered()` from the tidytext package.

As an example, I will present the countries with the lowest fertility rates across different years. I will retrieve the World Bank data with the `wbstats` package and add additional information with the `countrycode` package.

```{r}
library(wbstats)
library(countrycode)
library(tidyverse)
library(tidytext)
```

I start retrieving total fertility rate with the `wbstats::wb_data()` function.

```{r}
tfrt_data <- wb_data("SP.DYN.TFRT.IN", start_date = 2000, end_date = 2022) |>
  select(iso2c:SP.DYN.TFRT.IN)
```

Let's see the ten countries with lower fertility rates in years 2000, 2015 and 2022. We can do that using the `purrr::walk()`function.

```{r}

walk(c(2000, 2015, 2022), ~ {
  tfrt_data |>
  filter(date == .) |>
  arrange(SP.DYN.TFRT.IN) |>
  print(n = 10)
})
```

We observe that the ranking has been shifting during the three selected years. Let's retrieve the ten main countries in each year and save them in the `tfrt_shortlist` table.

```{r}
tfrt_shortlist <- tfrt_data |>
  filter(date %in% c(2000, 2015, 2022)) |>
  group_by(date) |>
  arrange(SP.DYN.TFRT.IN) |>
  slice(1:10) |>
  ungroup()
```

It can be interesting to add information about the continent of each country, using the `countrycode::countrycode()` package.

```{r}
tfrt_shortlist <- tfrt_shortlist |>
  mutate(continent = countrycode(iso2c, origin = "iso2c", destination = "continent"))
```

Let's try to do a vertical chart of fertility rate for each year using a faceted barplot.

```{r, out.width="100%"}
tfrt_shortlist |>
  mutate(country = fct_reorder(country, SP.DYN.TFRT.IN)) |>
  ggplot(aes(SP.DYN.TFRT.IN, country, fill = continent)) +
  geom_col() +
  facet_grid(date ~ .)
```

We observe that the plot has several flaws:

-   The order of each facet is the same for all years.
-   All involved countries appear in each facet, even if they do not appear in the ranking of each year.

To remedy this, we can use the functions `reoder_within()` and `scale_y_reordered()` from the `tidytext` package. The function `reorder_within()` replaces `forcats::fct_reorder()`.

```{r, out.width="100%"}
tfrt_shortlist |>
  mutate(country = reorder_within(country, -SP.DYN.TFRT.IN, date)) |>
  ggplot(aes(SP.DYN.TFRT.IN, country, fill = continent)) +
  geom_col() +
  scale_y_reordered() +
  facet_grid(date ~ ., scales = "free_y")
```

Once achieved the reordering within each facet, we can proceed to present the final plot.

```{r, out.width="100%"}
tfrt_shortlist |>
  mutate(country = reorder_within(country, -SP.DYN.TFRT.IN, date)) |>
  ggplot(aes(SP.DYN.TFRT.IN, country, fill = continent)) +
  geom_col() +
  scale_y_reordered() +
  facet_grid(date ~ ., scales = "free_y") +
  labs(title = "Evolution of lowest fertility rates", x = NULL, y = NULL) +
  theme_linedraw(base_size = 10) +
  theme(legend.position = "bottom", plot.title.position = "plot")
```

## Reference

-   World Bank Data. *Fertility rate, total (births per woman).* <https://data.worldbank.org/indicator/SP.DYN.TFRT.IN>

## Session Info

```{r, echo=FALSE}
sessionInfo()
```

