---
title: Structural Equation Modelling with lavaan
author: Jose M Sallan
date: '2023-01-10'
slug: -structural-equation-modelling-with-lavaan
categories:
  - R
  - statistics
tags:
  - dplyr
  - R
  - structural equation modelling
meta_img: images/image.png
description: Description for the page
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

In this post, I will introduce **structural equation modelling (SEM)**, a statistical technique to evaluate the fit of models including latent and observable variables taking as input the covariance matrix of observable variables.

SEM models can include **observable variables** (represented with squares in SEM diagrams) and **latent variables** (represented with circles in SEM diagrams). Latent variables represent constructs not observable directly. Observable variables are supposed to be measurements with error of constructs.

Any SEM model can be split into two submodels:

* **Measurement model**, specifying the relationship between observable and latent variables.
* **Structural model**, specifying the relationship between latent variables.

If the model you are examining consists solely of a measurement model, it is a **confirmatory factor analysis (CFA)** model. CFA is a special case of SEM, where we want to examine the fit of assigning observable variables to specific latent variables.

Let's see how can we fit SEM models in R. We will use the following packages:

* The `lavaan` package to fit SEM models.
* The `dplyr` package to handle data frames and `purr` to iterate effectively. Those packages belong to the tidyverse.
* The `broom` package to present a tidy output of `lavaan`.
* `semPlot` to plot diagrams of SEM models.
* `kableExtra` to present nice tables.

```{r}
library(lavaan)
library(dplyr)
library(purrr)
library(broom)
library(semPlot)
library(kableExtra)
```

To present the workflow I will use the `PoliticalDemocracy` dataset included in `lavaan`. A simpler workflow is presented in the `lavaan` documentation.

## Confirmatory factor analysis

Let's start examining a `m01` model where each observable variable is loading in the latent variable to which is attached theoretically. The latent-observable relationship is specified with a `=~`, with the latent variable on the left- and the observable variables in the right-hand side.

```{r}
m01 <- '
  # measurement model
    ind60 =~ x1 + x2 + x3
    dem60 =~ y1 + y2 + y3 + y4
    dem65 =~ y5 + y6 + y7 + y8'
```

Let's fit the `m01` model usign the `lavaan::cfa` function, specifying that we are using the `PoliticalDemocracy` dataset.

```{r}
f01 <- cfa(m01, data = PoliticalDemocracy)
```

We can use `f01` to obtain the path diagram with `semPlot::semPaths`:

```{r, out.width='80%', fig.align='center'}
semPaths(f01, edge.label.cex = 0.7, curvePivot = TRUE)
```

We can learn several things from this plot:

* Note that arrows go from latent to observable variables. This is contrary to general intuition, and means that all latent variables have a common source of variability, the latent variable, and a specific source of variability, introduced in the model as an error term.
* Some of the arrows are presented with a dashed line. This means that the regression coefficient of the observable variable model is set to one. This is done to scale the latent variable. 

Let's examine the output of the `lavaan` fit of the model. There are three ways of doing this:

* `summary` for a summary of selected results.
* `fitMeasures` for overall fit measures.
* `parameterEstimates` for a data frame of parameter estimates.

With `summary` we obtain a report of the model, including selected fit indices and parameter estimates:

```{r}
summary(f01)
```

With `fitMeasures` we obtain a long list of fit indices:

```{r}
fitMeasures(f01)
```

The output of `fitMeasures` is a named vector. If we want only some of the fit indices we can do as follows:

```{r}
fitMeasures(f01, c("npar", "chisq", "tli", "agfi", "rmsea"))
```

The `broom::glance` returns a selection of fit indices as a tibble. I use `dplyr::select` to obtain the selection of fit indices.

```{r}
glance(f01) |>
  select(npar, chisq, tli, agfi, rmsea)
```

Finally, with `parameterEstimates` we obtain a data frame of parameter estimates:

```{r}
parameterestimates(f01)
```

## Comparing CFA models

Let's compare the `m01` model with two other models:

* A `m00` model where all variables are loading in the same factor.
* A `m02` model where some of the observable variables are allowed to correlate, as specified in the original model. Covariances between variables are specified with a `~~` operator.

Comparisons between models similar to `m00` and `m01` are frequently used when reporting CFA analysis. If the confirmatory model works well, it will show a considerably better fit that the one-factor model.

```{r}
m00 <- '
  #measurement model
    f =~ x1 + x2 + x3 + y1 + y2 + y3 + y4 + y5 + y6 + y7 + y8'

m02 <- '
  # measurement model
    ind60 =~ x1 + x2 + x3
    dem60 =~ y1 + y2 + y3 + y4
    dem65 =~ y5 + y6 + y7 + y8
  # residual correlations
    y1 ~~ y5
    y2 ~~ y4 + y6
    y3 ~~ y7
    y4 ~~ y8
    y6 ~~ y8'
```

If we need to compare models, it can be a good idea to build an iterative workflow with `purrr::map`. Let's start storing the models in a list:

```{r}
cfa_models <- list(m00, m01, m02)
```

Let's fit the three models at once:

```{r}
fit_cfa_models <- map(cfa_models, ~ cfa(., data = PoliticalDemocracy))
```

We can use broom::glance to obtain the selected indices for each model as data frames, and purrr::map_dfr to put them all together in a data frame of three rows:

```{r}
fit_indices <- map_dfr(fit_cfa_models, ~ glance(.) |> select(npar, chisq, tli, agfi, rmsea))
```

And finally label each model with its name:

```{r}
fit_indices <- bind_cols(model = c("m00", "m01", "m02"), fit_indices)
```

Here are the results:

```{r}
fit_indices |>
  kbl() |>
  kable_styling(full_width = FALSE)
```

We observe that the fit of `m00` is much poorer than of the other two models. As the degrees of freedom of `m02` are smaller than `m01`, its fit is better. Let's examine the m02fitted model with `semPlot::semPaths`.

```{r, fig.align='center', out.width='100%'}
semPaths(fit_cfa_models[[3]], what = "path", whatLabels = "par", edge.label.cex = 0.7, curvePivot = TRUE)
```

## A structural model

Let's define the structural model presented in the lavaan website. The relationships of the structural model are represented by the `~` symbol.

```{r}
m03 <- '  
  # measurement model
    ind60 =~ x1 + x2 + x3
    dem60 =~ y1 + y2 + y3 + y4
    dem65 =~ y5 + y6 + y7 + y8
  # structural model
    dem60 ~ ind60
    dem65 ~ ind60 + dem60
  # residual correlations
    y1 ~~ y5
    y2 ~~ y4 + y6
    y3 ~~ y7
    y4 ~~ y8
    y6 ~~ y8'
```

The model can be fit with the `lavaan::sem` function:

```{r}
fit_m03 <- sem(m03, data = PoliticalDemocracy)
```

The diagram of this model is:

```{r, out.width='100%'}
semPaths(fit_m03, what = "path", whatLabels = "par", edge.label.cex = 0.7, curvePivot = TRUE)
```

We can use `summary`, `lavaan::fitMeasures` and `lavaan::parameterEstimates` to see the output as with the previous models. Instead of this, I will obtain fit indices in the same way as of previous models, and add the result to the table of CFA fit indices.

```{r}
df_fit_m03 <- bind_cols(model = "m03", glance(fit_m03) |> select(npar, chisq, tli, agfi, rmsea))

fit_indices <- bind_rows(fit_indices, df_fit_m03)

fit_indices |>
  kbl() |>
  kable_styling(full_width = FALSE)
```

We observe that the fit of `m02` and `m03` is exactly the same.


The parameter estimates are presented as a data table with `lavaan`, but `broom::tidy` presents them in a better way as a tibble:

```{r}
tidy(fit_m03)
```

If we want to check the significance of the coefficients of the structural model, we can do:

```{r}
tidy(fit_m03) |>
  filter(op == "~") |>
  select(term, estimate, std.error, p.value) |>
  kbl() |>
  kable_styling(full_width = FALSE)
```

Here we observe that all coefficients are significant for *p* < 0.001.

The `lavaan` package is one of the most common packages to fit structural equation models in R. Here I have presented how to retrieve fit indices and parameter estimates from a fitted model, and how to use the `broom` and `purrr` functionalities to fit several models at once.

## References

* The `lavaan` tutorial website: <https://lavaan.ugent.be/tutorial/>
* The `PoliticalDemocracy` example: <https://lavaan.ugent.be/tutorial/sem.html>
* The `semPlot` website examples section: <http://sachaepskamp.com/semPlot/examples>
* Mapping with `purrr`: <https://jmsallan.netlify.app/blog/mapping-with-purrr/>

## Session info

```{r, echo=FALSE}
sessionInfo()
```





