---
title: Plotting complex variable functions in R
author: Jose M Sallan
date: '2022-10-27'
slug: plotting-complex-variable-functions-in-r
categories:
  - R
tags:
  - ggplot
  - plotly
  - R
meta_img: images/image.png
description: Description for the page
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

In this post, I will present how to plot complex variable functions. To do so, I will plot the gamma function, as it allows illustrating how to plot asymptotic functions. I will plot the function in its real domain first, and then produce 2D and 3D plots of the complex domain. I will use `ggplot2` for two dimensional plots, and `plotly` for the 3D surface plot. To calculate the gamma function for complex numbers with imaginary component I will use the `pracma` package.

```{r}
library(tidyverse)
library(pracma)
library(plotly)
```

The **gamma function** is defined for any complex number $z$ as:

\[ \Gamma \left(z\right) = \int_0^\infty t^{z-1}e^{-t} dt \]

For every positive integer we have:

\[ \Gamma \left(n\right) = \left(n-1\right)! \]

That's why the gamma function is considered as a one of the possible extensions of the factorial for complex numbers.

We can plot any function straight away with ggplot using the `stat_function` geom. Let's see a straight away example:

```{r, fig.align='center', out.width='60%'}
ggplot() +
  stat_function(fun = \(x) x^2,
                xlim = c(-2, 2)) +
  theme_minimal()
```

In `stat_function`, the function to plot is entered in `fun`, and the function domain to plot as `xlim`.

## The gamma function for real values

Let's do something similar with the gamma function, using the R base function `gamma`, for a range of positive and negative numbers. Here I am adding more arguments to `stat_function`:

*  `n`, the number of points to use to draw the curve.
* `size` for line width. 

I have also added a `scale_x_continuous` to define the `breaks` of the x axis.

```{r, fig.align='center', out.width='80%'}
ggplot() +
  stat_function(fun = gamma, 
                xlim = c(-7, 7), 
                n = 1001,  
                size = 1.2) +
  scale_x_continuous(breaks = -4:2) +  
  theme_minimal()
```

The resulting plot is not satisfactory, as the behavior for non-negative numbers is quite different from the rest of the domain. There seems to be vertical asymptotes for nonpositive integer numbers, and the function grows factorially for positive values. Therefore, I have opted to present two plots for the gamma function:

* a plot for positive real values, showing the relationship between gamma and factorial.
* a plot for non-positive real values, showing the asymptotic behaviour of the function.

This is the plot for positive values:

```{r, fig.align='center', out.width='80%'}
df1 <- tibble(x = 1:8,
             y = factorial(x-1))

ggplot(df1, aes(x, y)) +
  geom_point(size = 2) +
  stat_function(fun = gamma, size = 0.8, color = "#A0A0A0") +
  theme_minimal() +
  scale_x_continuous(breaks = 1:8) +
  ggtitle("gamma function (positive values)")
```

The dots in the plot represent $\left(n-1\right)!$ for integer values. Its values come from table `df1`, and are plotted with `geom_point`. The gamma function is plotted with `stat_function`, adding a `color` argument. I have also added a title with `ggtitle`.

This is the plot for non-positive values.

```{r, fig.align='center', out.width='80%'}
df2 <- tibble(x = seq(-4, 2, length.out = 1000),
             y = gamma(x)) %>%
  mutate(y = ifelse(abs(y) < 20, y, NA))

ggplot(df2, aes(x, y)) +
  geom_line(size = 1.2) +
  geom_vline(xintercept = -4:0, size = 0.5, linetype = "dashed") +
  scale_x_continuous(breaks = -4:2) +
  theme_minimal() +
  ggtitle("gamma function (negative values)")
```

For the negative values, I do not use `stat_function`. Instead I defining values for the two axis in the `df2` table. When the absolute value of the function is greater than 20 I turn it no `NA`, so the `geom_line` is interrupted. Finally, I am adding asymptotes for non-positive integers using `geom_vline`. Note that the value of `xintercept` is a vector, so all asymptotes are plotted in a single instruction.

## The gamma function for complex values

Let's see now how we can learn how is behaving the gamma function for the whole complex domain, including values with nonzero imaginary component. R has a `cplx` class for complex numbers, with `real` and `imaginary` components. The polar components can be obtained with the `Mod` and `Arg` functions.

The gamma function for the whole complex domain can be obtained with the `gammaz` function of the `pracma` package. To represent the function graphically, I have chosen to assign the real and imaginary component to axis `x` and `y`, and the module of the gamma function to axis `z`. I nave used the expand_grid function to define a grid of points, so we have 10,000 points in a 100x100 grid.

```{r}
values <- expand_grid(r = seq(-2.5, 2.5, length.out = 100),
                      i = seq(-2.5, 2.5, length.out = 100)) %>%
  mutate(z = complex(real = r, imaginary = i),
         gamma = gammaz(z),
         mod = Mod(gamma),
         mod = ifelse(mod < 10, mod, 10))
```

We have two possibilities to represent a 3D function in 2D:

* **Countour lines:** we can do that with the `geom_contour()` with the function values in a `z` argument in `aes`.
* **Heatmap:** heatmaps are plot with `geom_tile()` with the function value in the `fill` argument in `aes`.

This plot presents a contour and a heatmap. I am plotting the tiles before the contour to present both, and I have added also a `alpha` value in the tile plot.

```{r, fig.align='center', out.width='80%'}
ggplot(values, aes(r, i)) +
  geom_tile(aes(fill = mod), alpha = 0.5) +
  scale_fill_gradient(name = "z",
                      low = "#FFFFCC",
                      high = "#FF0000") + 
  geom_contour(aes(z = mod), color = "black", bins = 50) +
  theme_minimal()
```

In the plot appear asymptotes in the integer negative values of the real axis.

To present a real 3D plot, the best option is to generate a surface in `plotly`.

```{r, fig.align='center', out.width='100%'}
n <- 100

x <- seq(-2.5, 2.5, length.out = n)
y <- seq(-3.5, 2.5, length.out = n)

t <- expand.grid(x = x, y = y)

t$mod <- Mod(gammaz(complex(real = t$x, imaginary = t$y)))
t$mod <- ifelse(t$mod < 5, t$mod, 5)

z <- matrix(t$mod, n, n)

plot_ly(x = x, 
        y = y, 
        z = z) |>
  layout(scene = list(xaxis = list(title = "r"),
                      yaxis = list(title = "i"),
                      zaxis = list(title = "mod"))) |>
  add_surface(colorscale = "Reds")
```

## Session info

```{r, echo=FALSE}
sessionInfo()
```

