---
title: Joining relational tables in dplyr
author: Jose M Sallan
date: '2022-04-04'
slug: joining-relational-tables-in-dplyr
categories:
  - R
tags:
  - dplyr
  - R
meta_img: images/image.png
description: Description for the page
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(kableExtra)
```

The `dplyr` package for tabular data handling includes a set of functions for joining data frames with a relational structure. By relational structure, I mean that there are columns in each table that can relate the contents of the two tables.

I will illustrate how these functions work using data from the World Bank. In addition to `dplyr`, I am using the `wbstats` package to retrieve those data directly from the web.

```{r}
library(dplyr)
library(wbstats)
```

**Mutating joins** (also called *merges*) return a data frame with the columns of two data frames *x* and *y*:

```{r, echo=FALSE, out.width='100%'}
url <- "https://d33wubrfki0l68.cloudfront.net/aeab386461820b029b7e7606ccff1286f623bae1/ef0d4/diagrams/join-venn.png"
knitr::include_graphics(url)
```

* **inner join**: returns all rows from *x* where there are matching values in *y*. If there are multiple matches between *x* and *y*, all combination of the matches are returned.

* **full (outer) join**: returns all rows from both *x* and *y*.  Where there are not matching values, returns NA for the ones missing.

* **left join**: returns all rows from **x**. Rows of *x* with no match in *y* will have NA values in the *y* columns. If there are multiple matches between *x* and *y*, all combination of the matches are returned.

* **right join**: returns all rows from **y**. Rows of *y* with no match in *x* will have `NA` values in the *x* columns. If there are multiple matches, all combinations are returned.

These four mutating joins are implemented in `dplyr` with the functions `inner_join`, `full_join`, `left_join` and `right_join`.

## Getting data from the World Bank

Let's use the `wbstats` package to retrieve values of two series:

* The `SP.POP.TOTL` indicator of total population <https://data.worldbank.org/indicator/SP.POP.TOTL>
* The `SP.DYN.TFRT.IN `of total fertility rate in rates per woman <https://data.worldbank.org/indicator/SP.DYN.TFRT.IN>

```{r}
pop_wb <- wb_data("SP.POP.TOTL", start_date = 2000, end_date = 2022)
pop_wb
```

```{r}
tfrt_wb <- wb_data("SP.DYN.TFRT.IN", start_date = 2000, end_date = 2022)
tfrt_wb
```

I will edit the resulting tables removing missing values, retaining the columns with `country`, `date` and value and renaming the value.

```{r}
pop <- pop_wb %>%
  filter(!is.na(SP.POP.TOTL)) %>%
  select(country, date, SP.POP.TOTL) %>%
  rename(population = SP.POP.TOTL)

tfrt <- tfrt_wb %>%
  filter(!is.na(SP.DYN.TFRT.IN)) %>%
  select(country, date, SP.DYN.TFRT.IN) %>%
  rename(fertility = SP.DYN.TFRT.IN)
```

## Joins by one variable

I will start joining tables with data from year 2010:

```{r}
pop_2010 <- pop %>%
  filter(date == 2010) %>%
  select(-date)

tfrt_2010 <- tfrt %>%
  filter(date == 2010) %>%
  select(-date)
```

Both tables have different number of observations:

* `pop_2010` has `r nrow(pop_2010)` rows.
* `tfrt_2010` has `r nrow(tfrt_2010)` rows.

It is important to track the number of rows of each table to examine the effect of each join. Let's begin with the inner join. Like in the rest of joins, I have declared the joining columns with `by`. If no value of by is passed, the functions pick the columns of equal name in both tables.

```{r}
inner_2010 <- inner_join(pop_2010, 
                         tfrt_2010, 
                         by = "country")
inner_2010
```

The inner join will have the least rows of all joins, as it only includes values of the joining columns present in both tables. It has `r nrow(inner_2010)` rows, the same as `tfrt_2010`.

```{r}
full_2010 <- full_join(pop_2010, 
                       tfrt_2010, 
                       by = "country")
full_2010
```

The full join has the maximum value of rows of all joins, as it includes the observations included in any of the tables. It has `r nrow(inner_2010)` rows, the same as `pop_2010`.

With the full join, we can examine which rows do not include fertility or population:

```{r}
full_2010 %>%
  filter(is.na(fertility))
full_2010 %>%
  filter(is.na(population))
```

We observe that some countries that have populations have not fertility rate records, but not in reverse.

Left and rigth joins will have the same rows of the left and right tables in the join expression, respectively.

```{r}
left_2010 <- left_join(pop_2010, 
                       tfrt_2010, 
                       by = "country")
left_2010
```

```{r}
right_2010 <- right_join(pop_2010, 
                         tfrt_2010, 
                         by = "country")
right_2010
```

## Joins by two variables

Let's join the two original `pop` and `tfrt` tables. We need to do the join by two variables: `country` and `year`.

```{r}
inner <- inner_join(pop, 
                    tfrt, 
                    by = c("country", "date"))

full <- full_join(pop, 
                  tfrt, 
                  by = c("country", "date")) 

left <- left_join(pop, 
                  tfrt, 
                  by = c("country", "date"))

right <- right_join(pop, 
                    tfrt, 
                    by = c("country", "date"))
```

Let's see how many rows we have in each join and the two original tables:

```{r, echo=FALSE}
rows_df <- data.frame(join = c("pop", "tfrt", "inner", "full", "left", "right"),
                      rows = sapply(list(pop, tfrt, inner, full, left, right), nrow))
names(rows_df)  <- c("table", "number of rows")

rows_df %>%
  kbl() %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center")
```

Now the `inner` table has less rows than `tfrt`, while the `full` table has more rows than `pop`. This means that there are observations in `pop` missing in `tfrt`, and vice versa.

## Filtering join

To check which observations of one table are missing in the other, we can use the `anti_join` filtering join: `anti_join(x, y)` returns observations of `x` with no match in `y`.

We can see which elements of `pop` are not present in `tfrt` doing:

```{r}
anti_join(pop, tfrt, by = c("country", "date"))
```

We have `r nrow(anti_join(pop, tfrt, by = c("country", "date")))` observations in this table. This is not surprising, as fertility records and not as exhaustive as population's.

Let's examine which elements of `tfrt` are not present in `pop`:

```{r}
anti_join(tfrt, pop, by = c("country", "date"))
```

Here we have only `r nrow(anti_join(tfrt, pop, by = c("country", "date")))` observations. None of them is from year 2010, so we did not observed them in the previous section.

`dplyr` offers a set of functions to join relational tables that allow performing **mutating joins**, because they return a new table with information from the two inputs. This allows doing SQL-like operations inside the R environment. The four mutating joins (inner, full, left and right) are a replacement for the R base `merge` function, hopefully in a more intuitive fashion. **Filtering joins** allow controlling for which rows have no match in the other table, so they do not return a new table.

## References

* *Filtering joins.* <https://dplyr.tidyverse.org/reference/filter-joins.html>
* *Mutating joins.* <https://dplyr.tidyverse.org/reference/mutate-joins.html>
* *`wbstats`: An R package for searching and downloading data from the World Bank API.* <https://cran.r-project.org/web/packages/wbstats/vignettes/wbstats.html>

## Session info

```{r, echo=FALSE}
sessionInfo()
```

