---
title: Using Recursion to Generate Feature Interaction Terms
author: Jose M Sallan
date: '2024-05-31'
slug: using-recursion-to-generate-feature-interaction-terms
categories:
  - R
tags:
  - machine learning
  - R
  - R programming
meta_img: images/image.png
description: Description for the page
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In the context of prediction jobs, we can be interested in generating products of powers of variables. These products can be used as features for regression-based prediction techniques like neural networks.

For small values of $p$ and $n$, these interactions are straightforward. For instance, for $p=2$ and $n=2$ these are:

-   $x_1$
-   $x_2$
-   $x_1^2$
-   $x_2^2$
-   $x_1x_2$

For larger values of $p$ and $n$ we may need a systematic way of generating those products of powers. Here I will suggest a recursion-based procedure to generate all products of up to $p$ powers of a set of $n$ variables and provide an implementation in R base. I will be using this result to obtain the interaction terms for a set of $n$ features whose sum of powers is equal to $n$.

## Recursion

As defined in Wikipedia, recursion occurs when the definition of a concept or process depends on a simpler or previous version of itself. To define a recursion we need:

-   A set of base cases that do not need recursion to produce an answer. In the Fibonacci series, these are `fibonacci(0) = 0` and `fibonacci(1) = 1`.
-   A recursive step that reduces all the other cases to the base cases. In the Fibonacci series, the step is `fibonacci(n) = fibonacci(n-1) + fibonacci(n-2)`.

In R base the implementation of recursion is straightforward, using a function covering the base cases and the recursive step. Note that the function appears within the function definition.

```{r}
fibonacci <- function(n) {
  if (n <= 1) {  # Base case: If n is 0 or 1, return n
    return(n)
  } else {
    # Recursive case: Calculate the nth term using recursion
    return(fibonacci(n - 1) + fibonacci(n - 2))
  }
}
```

Some values of the function

```{r}
fibonacci(0)
fibonacci(10)
```

A sequence of values of the Fibonacci series:

```{r}
sapply(0:20, fibonacci)
```

## Products of Powers

To define the products of powers, we will represent them as vectors of length $n$ with values from 0 to $p$. The value zero contemplates the case where a variable is not included in the product. For $n = 3$ and $p = 2$ the product $x_1x_3^2$ is encoded as $(1, 0, 2)$.

In the suggested recursive process, we are adding a variable at each step. The base case is for $n=1$. If we have $p=3$, this base case is:

```{r}
s1 <- matrix(0:3, 4, 1)
```

For $n=2$ we will use the matrix of $n=1$:

```{r}
s2 <- lapply(0:3, function(i) cbind(s1, i))
s2
```

We can bind the rows together doing:

```{r}
do.call(rbind, s2)
```

For $n$ variables, we need to implement this process recursively:

```{r}
powers <- function(p, n){
  if(n == 1){
    s0 <- matrix(0:p, p + 1, 1)
    return(s0)
  }else{
    sn <- lapply(0:p, function(i) cbind(powers(p, n-1), i))
    m <- do.call(rbind, sn)
    colnames(m) <- NULL
    return(m)
  }
}
```

Let's see some examples:

```{r}
powers(3, 1)
powers(3, 2)
powers(3, 3)
```

The function generates $p^n$ powered product terms.

## Interactions of order *n*

Of all generated powered products, we are interested in picking only those whose sum of exponents are equal to $n$. For $p =2$ and $n = 2$ these are:

-   $x_1^2$
-   $x_2^2$
-   $x_1x_2$

So we need to exclude $x_1$, $x_2$ and the intercept $1$ from this list. We do that generating all `power(n, n)` interactions and filtering them adequately.

```{r}
interactions <- function(n){
  
  # all powered products
  t <- powers(n, n)
  
  # rows summing n
  r <- apply(t, 1, \(x) sum(x) == n)
  
  # selectin rows summing n
  t_f <- t[r, ]
  return(t_f)
}
```

Then we have:

```{r}
interactions(2)
```

For larger values we have results like:

```{r}
interactions(4)
```

## References

-   Recursion: <https://en.wikipedia.org/wiki/Recursion>
-   Combine a list of matrices to a single matrix by rows <https://stackoverflow.com/questions/16110553/combine-a-list-of-matrices-to-a-single-matrix-by-rows>

## Session Info

```{r, echo=FALSE}
sessionInfo()
```



